Image Correlation Spectroscopy is a Robust Tool to Quantify Cellular DNA Damage Response

Version 1

Abstract

Abstract

The DNA Damage response (DDR) is both essential and highly complex. Evaluating the DDR is a critical aspect of cell biology. Counting DNA damage foci is one of the most common approaches to study the DDR. Yet, quantification of protein foci suffers from experimental limitations, subjectivity of analysis and is restricted to a handful of the hundreds of DDR proteins. Here we apply image correlation spectroscopy (ICS) to quantify the local clustering at sites of DNA damage directly. We found that ICS outperformed foci counting of traditional DDR markers and enabled quantification of other markers without the complex labeling procedures that are otherwise required. ICS analysis also provided insight into DDR protein recruitment that was previously undetectable. Further expansion incorporating analysis to cell cycle classification demonstrates a rapid, non-biased approach to fully study the DNA damage response within cells. ICS analysis presents an objective, quantitative image analysis technique to study the DNA damage response in unaltered cells that we expect will significantly enhance quantitative DNA damage response research.

Introduction

The DNA damage response (DDR) represents multiple different pathways that are dependent on several factors including cell cycle and the type of damage1–3. DNA damage and the DDR are prominent in several disease areas such as cancer4, neurodegeneration5 and aging6, and the aberrant DDR is a major target of cancer therapies7. Collectively, the DDR comprises hundreds of proteins that are recruited to sites of damage in a concerted manner serving as scaffolds, DNA binders, modifying enzymes and repair enzymes. Largely due to the massive evolutionary pressure to maintain the genome, the DDR is enormously complex, rendering it difficult to study in cells.

Current methods to evaluate DNA damage vary from molecular strategies such as polymerase chain reaction8, comet assays9 and gel electrophoresis10 to intracellular approaches that include, fluorescence in situ hybridization11, radioimmunoassay12 and TUNEL assays13. Analytical methods have also been employed, for example, gas or liquid chromatography, mass spectrometry and electrochemical approaches have been used to identify DDR components and signaling7. However, the gold standard for assessing DNA damage levels in cells is fluorescence imaging and counting DNA repair foci formation by labeling established protein markers, such as Ku70/8014, γH2AX15 or XRCC116, through immunofluorescence or genetically encoded fluorescent proteins.

Historically, foci counting was performed manually and relied on individual researchers to identify foci, introducing an inherent bias to the process. More recently, foci counting has been widely automated via computer-based algorithms17,18. Numerous software tools have been released to count foci via methods of signal thresholding, maxima finding, and morphological reconstructions to accurately detect foci in nuclei. However, foci are not often individually distinguishable and are present at varying sizes and intensities, leading to some ambiguity in thresholding, segmenting and counting parameters, with a wide variety of software employing different algorithms19–21. Furthermore, γH2AX foci were recently revealed to represent clustered structures of γH2AX nano-domains22, suggesting multiple sites of DNA damage are present in single foci. Thus, it is likely that not all foci are the same – larger foci representing a greater degree of or more mature DNA damage than smaller foci.

In addition to measurements of known foci markers, studies using laser micro-irradiation have shown that hundreds of different proteins are recruited to sites of DNA damage23, despite not forming foci that are resolvable in fluorescence images when more biologically relevant DNA damage is induced. Therefore, the role of the majority of DDR proteins remains poorly understood because their recruitment to DNA damage foci is difficult to quantify with traditional approaches. Typically, non-recruited protein is pre-extracted, through lysis of the cell and nucleus, prior to fixation to reduce the background concentration and enhance the resolution of protein foci. However, the demonstration that DNA damage foci are biomolecular condensates where foci protein concentration exists in equilibrium with the nuclear concentration24 suggests that pre-extraction may impact the DNA damage foci as well to hinder the accuracy of the results.

To overcome the limitations of current cellular DDR analysis approaches, we hypothesized that measuring clustering of labeled proteins within the nucleus could quantify the degree of DNA damage and involvement of any DDR protein. To test this approach, we employed fluorescence microscopy imaging with fluorescence fluctuation analysis via spatial image correlation spectroscopy (ICS) as a tool for measuring the clustering of proteins during DDR in multiple cancer cell lines treated with DNA damaging agents and DDR therapeutics. Fluorescence correlation spectroscopy (FCS), which requires temporal imaging, has been previously used for the measurements of macromolecular aggregation via autocorrelation analysis of fluorescence intensity fluctuations that arise due to changes in the local concentration of fluorescence molecules in a tiny focal volume25–27. Thus, FCS is well suited to study single, well-defined foci. Conversely, ICS was developed as the imaging analog of FCS to measure densities and aggregation states of membrane receptors from spatial correlation analysis of very small fluorescence fluctuations in single images28,29. Thus, ICS presents a potential alternative method to quantify DNA damage and expand the detection of DDR response elements beyond established protein markers.

Methods

Theory

Image correlation spectroscopy is based on the analysis of fluorescence intensity fluctuations arising from variations in the number of fluorescent particles within focal spots imaged in space using a fluorescence microscope. Under the assumptions of particle ideality and linear fluorescence emission (i.e., no signal saturation or energy transfer), the detected mean fluorescence intensity of tagged molecules varies linearly with the concentration of fluorophores in a focal spot (Fig. 2A-B). The square relative intensity fluctuation (intensity variance/mean intensity, Eq. 1) will be the mean number of detected independent fluorescent particles per focal spot, since ideal behavior entails that the molecules obey Poisson statistics within the volume: given intensity fluctuations defined as δi = i − 〈i〉, where 〈i〉 is the mean intensity and 〈n〉 is the mean number of independent fluorescent particles per focal spot area.

To obtain this number density information independent of white noise sources (such as shot noise), ICS calculates the mean intensity normalized spatial autocorrelation function of fluorescence intensity fluctuations (Eq. 2) of a region of interest (ROI). The orthogonal spatial lag variables ξ and η represent discrete pixel shifts in x and y directions in an image at which the spatial correlation is calculated. The zero spatial lags value of the spatial autocorrelation function is the square relative intensity and hence the particle number density.

For computational speed, the spatial correlation function is calculated using Fourier methods (Eq. 3)28, where F is the discrete 2D spatial fast Fourier transform of the ROI, F* is the complex conjugate and F−1 is the inverse Fourier transform.

The calculated spatial intensity fluctuation correlation functions are then each fit to a 2D Gaussian (Eq. 4) using a non-linear least-squares algorithm, where the zero lags point is not weighted due to white noise contributions. Output fit parameters are g(0, 0), the zero-lags amplitude, ω0, the e−2 Gaussian correlation radius, and g∞, the long spatial lag offset.

The best fit zero lags amplitude of the correlation function is an estimate of the square relative fluctuation from Eq. 1, with an inverse that is the mean number of independent fluorescent particles per focal spot area, 〈n〉 (Eq. 5). The total number of particles (TNoP) in the ROI can be calculated using the image area and ω0 (Eq. 6).

When fluorescent molecules undergo aggregation or clustering, the clusters/aggregates are not resolvable using diffraction limited optical microscopy (i.e., confocal) unless large aggregates are present (such as individually distinguishable foci). However, aggregation manifests in the ROI as larger relative intensity fluctuations from a smaller number of brighter fluorescent particles per focal spot. If the number of overall fluorophores does not change, the mean fluorescent intensity should be constant, but the number of independent fluorescent particles (formed of one or more fluorophores) should decrease after clustering (Fig. 2 B-C). Since the average intensity of the ROI is proportional to the total number of fluorophores 〈nf〉 after background correction and a constant (c) relating intensity to fluorophore count, a degree of aggregation (DA) measurement can be calculated using the average intensity, 〈i〉, and the mean number of fluorescent particles, 〈n〉; To first order, the degree of aggregation is proportional to the mean number of fluorophores per aggregate if the variance of the aggregate distribution is small29.

Image Analysis

For the ICS analysis, nuclei were segmented from the DAPI channel using a watershed approach in MATLAB. Briefly, we applied a difference of gaussians for size-based feature extraction of nuclei in the image (upper gaussian σ=100, lower gaussian σ=2-5 dependent on cell size). This feature enhanced image was then binarized to create a primary mask (using the function imbinarize) and connected components under 100 pixels were removed (using the function bwareaopen). We then created a distance transform of the resulting binary image (using the function bwdist). Seed points were determined by finding the regional maximum of a gaussian filtered image, with a gaussian σ of appropriate size for seed point detection (σ=7 or 8 dependent on cell size). The seed points were applied to the distance transform, and we applied the watershed function to the resulting image to obtain masks for each nucleus (extracted using the function regionprops).

Using the nucleus masks, we performed ICS analysis on the antibody-stained channel, within size limitations for a nucleus (objects > 400 and < 2400 pixels). Intensity, DA and TNoP were calculated for each cell. In this analysis, since the ROI is defined by the nucleus mask (segmented from nuclear signal using DAPI staining), TNoP represents the total number of particles per nucleus, with the image area being the number of pixels in each nuclear mask.

Measurements across experiments were normalized by the control in each dataset. The confocal imaging settings were kept the same across each dataset to provide comparable measurements per dataset. Data were plotted in Prism and statistical analysis was performed in Excel.

Local maxima were used to count foci in ImageJ/FIJI, using the tool Find Maxima under the Process menu30. This tool requires a threshold input parameter (Prominence/Tolerance), where a threshold is set at the maximum value minus noise tolerance and local maxima must be greater than this threshold to be counted. To compare the effect of the input parameter, we performed measurements at Tolerance = 300 and 600. The final measurement was taken at the user-defined threshold where the user visually inspected different Tolerance values and chose a value that visually removed background (thresholds between 300 and 600) – a subjective process.

Imaris spots analysis were performed with Imaris 9.8 with background subtraction, using a spot size of 1 µm that was measured from the smallest distinguishable foci from the control image. For images with multiple nuclei, ImarisCell was used with the detection type “cell and vesicles”, where the nucleus in the DAPI channel was modelled as the “cell” and antibody-labeled foci were modelled as the “vesicles”. This enabled us to count the number of foci per nucleus for multiple cells. In each dataset, the automatic “quality” threshold was determined from the control image and the same threshold was used for every image in the dataset to be able to compare the outputs across datasets. Similar to the ICS analysis, only cells with an area between 400-2400 voxels were included for analysis.

Materials and Cell Culture

All chemicals were purchased from Sigma unless otherwise noted. All drugs were purchased from Selleck Chemicals. Cells were cultured in media supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin. OVCA429 were originally obtained from Dr. Michael Birrer, HT1080 were obtained from ATCC cultured in DMEM. SKOV3, HCC1937, HCC1395, MCF7, UWB1.289 and UWB1.289 +BRCA1 were obtained from ATCC and cultured in RPMI. UWB1.289res were made resistant to olaparib through long-term culture in up to 1 μM olaparib. Unless otherwise indicated, for MMS treatment, cells were incubated for 6 hours before fixation with 4% formaldehyde, whereas for treatment with PARP inhibitors and other drugs, cells were incubated for 24 hours before fixation.

Cell Dose Dependence Response

Cells were plated in a 96-well plate at 2×103 cells per well and treated with drug for 5 days in five replicate wells. Cell viability was then determined using Presto Blue (Thermo Fisher). Signal was averaged over replicate wells and normalized to blank (no cells) as well as to the levels of untreated wells. A sigmoidal response curve was fit to the average of two experiments in Prism.

Immunofluorescence and Fluorescence Microscopy

Cells were plated in 8-well slides (Fisher) or 384 well glass bottom plates (CellVis) to achieve 50% confluence overnight. Cells were then treated and fixed with 4% formaldehyde in PBS at room temperature for 10 minutes. Antibody staining was performed according to the manufacturers protocol (Cell Signaling Technologies or Abcam). Primary antibodies used were γH2AX (CST, #2577), RPA1 (CST, #2267), RAD51 (Abcam, 63801) and PCNA (CST, #2586), with Anti-rabbit IgG Fab2 Alexa Fluor 647 (CST, #4414) as the secondary fluorescent antibody. After antibody labeling, prolong gold with DAPI (CST) or DAPI (1 μg/ml) in PBS for well plate experiments was added and samples were stored for up to 1 week prior to imaging. Confocal imaging was performed on an Olympus FV1000 multiphoton/confocal microscope using an Olympus XLUMPlanFL N 20x objective, NA 1.00 with chromatic correction, pinhole at 1 Airy unit and 3x digital zoom. DAPI was excited with a 405 nm laser, while Alexa Fluor 647 was excited with a 635 nm laser. A multiband (DM 405/488/561/633 nm) dichroic mirror was used to direct the excitation laser and collect emission. Emission was then separated into two channels through a dichroic mirror (490nm) and DAPI signal was cleaned up by diffraction grating, which is the low band pass filter used by this instrument. Widefield imaging were performed on a Nikon Ti2 with a 20x (NA 0.75) objective using standard DAPI and Cy5 filter cubes.

Western Blotting

Cells were lysed with RIPA buffer (Thermo Fisher) containing protease inhibitors (Thermo Fisher), vigorously vortexed and incubated on ice for 15 minutes. Lysate was then pelleted by centrifugation at 14,000g for 15 min at 4 °C. Supernatant was transferred to a clean tube and total protein was determined using a BCA assay (Pierce). Protein was loaded into a NuPAGE gel (Thermo Fisher) and transferred to nitrocellulose paper (Thermo Fisher). Blocking and antibody labeling was then carried out according to the antibody manufacturer’s protocol. GAPDH (R&D systems, #AF5718) was used as a loading control.

Statistics and Reproducibility

All experiments were performed in multiple repeats as indicated in the figure legends. Statistical significance was calculating using one way ANOVA or Students’ t test.

Results

Counting Foci via Fluctuation Analysis

Most foci counting methods are based on finding the local fluorescence intensity maxima either through intensity thresholding to create masks of high intensity single foci or implementing local maxima finding algorithms. Here, we evaluated traditional local maxima finding algorithms on the ovarian cancer cell line SKOV3 by creating resolvable DNA damage with the PARP inhibitor olaparib (10 µM) and antibody-labeling γH2AX to visualize foci (Fig. 1B). To enable clean isolation of nuclei, the cells were also labeled with DAPI (Fig. 1A). First, we used the Find Maxima tool in ImageJ/FIJI which detects local maxima to define individual foci. However, this approach relies heavily on the intensity threshold parameter (Prominence/Tolerance) used to remove the background signal. This parameter is defined by the user and is highly subjective – a high value causes some foci to be omitted from the count, yet, conversely, a low value leads to background signal being counted as foci. Therefore, we evaluated images at 2 different intensity thresholds (Fig. 1D-E) in 21 single cells to determine the ambiguity of the threshold parameter (Fig. 1G) before manually determining a threshold for each cell for the final measurement (Fig. 1H).

Common approaches to count DNA damage foci.

(A) DAPI channel with 5 µm scale bar, (B) γH2AX channel, (C) Composite of A) and B). (D-E) Foci counting with Find Maxima tool in ImageJ with a threshold of D) 600 and E) 300. (F) Foci counting using Imaris Spots with an automatic threshold and an expected spot size of 1 µm. (G) Number of maxima per nucleus from different thresholds using ImageJ for the same cell. (H) Foci counting comparison between a user defined threshold in ImageJ and automatic threshold in Imaris on the same cell. (I) Intensity adjusted γH2AX channel showing lower intensity foci (white arrows), larger foci (green arrows) and higher background intensities (blue arrows).

In addition, we implemented a different local maximum finding algorithm with automatic thresholding using Imaris Spots detection (Fig. 1F). Spots detection uses a Gaussian filter to smooth the image and foci are then located at the local maxima of the filtered image. Automatic thresholding is based on a k-means statistical method and Spots uses an expected foci size to identify multiple spots in a large cluster where individual foci are indistinguishable and could be classified as single foci with local maxima detection. Our results produce foci counts that are dependent on the approach used and setting implemented (Fig. 1 and S1). In our analysis, different thresholds/ algorithms give different counts for the same cell due to different intensity foci (white arrows, Fig. 1I), different size foci (green arrows, Fig. 1I) and non-uniform background intensities (blue arrows, Fig. 1I).

We next employed fluctuation analysis via spatial ICS in the same set of images. Uniquely ICS analyzes the entire dynamic range of detected fluorescence, not just intensity peaks that could be found with local maxima algorithms. We employed ICS analysis on segmented nuclei (Fig. 2D) and calculated the total number of particles (TNoP) parameter per nucleus for comparison to foci counting data from ImageJ/FIJI and Imaris. Direct, cell by cell comparison of ICS TNoP with both the user-optimized threshold ImageJ Find Maxima (Fig. 2E) and Imaris Spots (Fig. 2F) foci counting results produced a strong correlation. Here, ICS is determining the total number of independent particles, leading to a TNoP value that is not directly counting resolvable foci. However, the correlation to the other approaches demonstates that ICS is measuring traditional foci forming DNA damage.

Counting foci using image correlation spectroscopy correlates with optimized measurements using existing approaches.

(A) Schematic population of monomeric fluorescent particles where emitting particles (green) are in a diffraction limited excitation focus (cyan) – not to scale. (B) Schematic showing increase in particle density from A) which results in an increase in average fluorescence intensity. (C) Schematic showing clustering of fluorescent particles where the total number of independent particle clusters is the same as A). Spatial ICS measures the average number of independent fluorescent units per area which, along with the average fluorescence intensity, is used to calculate the average degree of aggregation of the particles. (D) Intensity-based segmentation of the DAPI channel with an automatic threshold using the MATLAB function imbinarize to create a mask of the nucelus. The mask specifies the ROI for ICS analysis of the γH2AX channel, outputting mean number of independent fluorescent particles per focal spot area (inverse of the intensity normalized spatial correlation function amplitude). (E) and (F) Comparison of the ICS parameter TNoP with E) ImageJ Find Maxima tool with a user defined threshold and F) Imaris Spots with an automatic threshold for the same nuclei. A linear line of best fit is shown on each graph with the resulting R2.

Particle Counting versus Aggregation in γH2AX signal

One major goal of foci counting in cells is to determine the degree of DNA damage. Therefore, after validating our foci counting methods at the individual cell level, we sought to evaluate ICS as a tool to quantify the DDR response in bulk populations of cells. We first treated SKOV3 cells with varying concentrations of methyl methane-sulfonate (MMS) or the PARP inhibitors veliparib or olaparib and evaluated DNA damage through immunofluorescence of γH2AX. MMS is a DNA alkylating agent that directly damages DNA by modifying guanine and adenine to instigate mispairing of bases and replication blocks31. PARP inhibitors, however, do not necessarily directly induce DNA damage, but stall the DNA damage response by “trapping” PARP1 and preventing recruitment of critical DNA damage response proteins23,32–34. Therefore, these two drug classes should have different impacts on foci formation and foci composition.

MMS treatment produces a wide range of γH2AX intensity and visible foci across individual cells (Fig. 3A-C; 0.1 mM MMS and Fig. S2; 0 mM and 1 mM MMS). Thus, analysis capable of detecting both low intensity foci and overlapping, highly abundant foci within a single experiment is needed for accurate measurements over a range of conditions. In the non-contrast adjusted γH2AX image (Fig. 3B), brighter γH2AX foci are clearly distinguishable, however once the image is contrast adjusted (Fig. 3C), lower intensity foci are apparent and the higher intensity foci appear as large foci clusters, especially in nuclei demonstrating more DNA damage. To overcome the inherent distribution of DNA damage across cells, we scaled up the analysis to examine hundreds of cells per image by acquiring large, stitched images (Fig. S3). Here, due to the dependency of selecting the proper background intensity per cell, we omitted the ImageJ Find Maxima analysis and analyzed data using ICS and Imaris Spots. To directly compare analyses, each measurement within an experiment was normalized to the average value of DMSO treated control cells.

Imaris Spots and ICS capture dose dependent DNA damage through γH2AX foci in SKOV3 cells.

(A-C) SKOV3 cells treated with 0.1 mM MMS; A) DAPI channel with 10 µm scale bar, B) γH2AX channel, C) γH2AX channel with intensity-adjustment. (D-L) Data for normalized intensity, normalized number of spots and normalized ICS DA of γH2AX signal across different MMS (D-F), veliparib (G-I) and olaparib (J-L) treatment concentrations in SKOV3 cells. Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. MMS; N=2423-3160, veliparib; N=1818-2922, olaparib; N=1489-2956. * indicates p<0.05, ** indicates p<0.001, *** indicates p<1X10−10 via one-way ANOVA significance testing.

In analyzing DNA damage, the amount of γH2AX per cell is represented by the normalized integrated fluorescence intensity (Fig. 3). We found that γH2AX intensity experienced small, significant increases at lower MMS doses but showed a massive (nearly 8X), significant increase at the highest dose, results that agree with SKOV3 sensitivity to MMS (Fig. S4). PARP inhibitor treatment produced similar results except for a slight decrease at 1 µM veliparib. The number of spots determined by Imaris captures the dose dependent DNA damage induced by MMS and PARP inhibitors as expected (Fig. 3), with the response to olaparib plateauing after 10 µM. Yet, surprisingly, at the highest dose of PARP inhibitor, the number of foci was higher under veliparib treatment, the weaker PARP inhibitor (Fig. S4).

We then analyzed the same images with ICS. Microscopic number density calculated by ICS often decreases in cases of significant aggregation, resulting in a smaller number of more densely packed clusters35,36, therefore we calculated the ICS degree of aggregation (DA) as a measurement that captures both low and high DNA damage. Critically, ICS DA is not a direct measurement of the number of foci, but rather a quantification of overall label clustering. DA results demonstrated both MMS and PARP inhibitor dose dependence (Fig. 3), similar to results obtained with Imaris Spots. Like Spots analysis, the ICS DA analysis of 100 µM veliparib produced a higher value than 100 µM olaparib, compared to their respective controls. Here, ICS DA proved comparable to Imaris Spots.

We next analyzed DNA damage dose dependence in another cell line, OVCA429. Fig. 4A-C displays OVCA429 cells exposed to 0.1 mM MMS, highlighting a range of γH2AX signal, spanning from only low diffuse signal in the nucleus, to single distinguishable γH2AX foci in the diffuse signal, to large clusters of different sizes in high intensity nuclei (Fig. 4C). γH2AX intensity significantly increases at 1 mM MMS (Fig. 4D) and Spots analysis captured the MMS dose dependent DNA damage (Fig. 4E) and ICS DA produced results similar to Spots analysis (Fig. 4F).

Imaris Spots and ICS DA capture dose dependent DNA damage through γH2AX in OVCA429 cells.

A-C) OVCA429 cells treated with 0.1 mM MMS; A) DAPI channel with 10 µm scale bar, B) γH2AX channel, C) γH2AX channel with intensity-adjustment. (D-L) Data for normalized intensity, normalized number of spots and normalized ICS DA of γH2AX signal across different MMS (D-F), veliparib (G-I) and olaparib (J-L) treatment concentrations in OVCA429 cells. Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. MMS; N=1194-2068, veliparib; N=1256-1997, olaparib; N=1445-2334. * indicates p<0.05, ** indicates p<0.001, *** indicates p<1X10−10 via one-way ANOVA significance testing.

Contrary to SKOV3 analysis, despite similar sensitivity to olaparib and veliparib (Fig. S4), γH2AX intensity in OVC429 in response to PARP inhibitors showed a lower change in signal with no significant change in intensity at higher concentrations (Fig. 4). However, Imaris Spots captured a significant dose-dependent response with all but 10 μM veliparib generating a significant increase from the next lower dose (Fig. 4). Thus, despite no changes in intensity, more foci are being formed at higher doses. ICS analysis, however, was more sensitive than Imaris Spots, capturing a significant dose-dependent change at each concentration (Fig. 4). Overall, in OVCA429, ICS DA produced similar results to Imaris Spots in olaparib treated cells, but measured dose dependent response in cells treated with veliparib where Imaris Spots was unable to capture the dose dependency.

ICS Captures Clustering of DNA Damage Proteins that do not form Resolvable Foci

Next, we sought to quantify the behavior of proteins that do not form resolvable foci during DNA damage without immunofluorescence pre-extraction, specifically Replication Protein A1 (RPA1) and RAD51. The RPA complex is required for major DNA repair pathways and modulates RAD51 recruitment to sites of DNA damage37,38. However, increased RPA1 and RAD51 concentrations created at sites of DNA repair are not always resolvable as individual foci38,39. In the absence of induced DNA damage, RPA1 immunofluorescence produces a diffuse signal throughout the nuclei in SKOV3 cells (Fig. 5). Only with contrast adjusting (Fig. 5C), does heterogeneous distribution of RPA1 become visible.

Dose-dependent DNA damage impact on RPA1 and RAD51.

(A-C) Control SKOV3 cells, (D-F) SKOV3 cells treated with 1 mM MMS; (A&D) DAPI channel with 10 µm scale bar, (B&E) RPA1 channel, (C&F) RPA1 channel with intensity-adjustment. Normalized RPA1 Imaris Spots (G) and normalized ICS DA (H) in SKOV3 and OVCA429 cells (I&J). Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. N=1600-1872 for SKOV3 and N=1320-2198 for OVCA429. (K-M) Control SKOV3 cells, (N-P) SKOV3 cells treated with 1 mM MMS; (K&N) DAPI channel with 10 µm scale bar, (L&O) RAD51 channel, (M&P) RAD51 channel with intensity-adjustment. Normalized RAD51 Imaris Spots (Q) and normalized ICS DA (R) in SKOV3 and OVCA429 cells (S&T). Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. N=1597-2241 for SKOV3 and N=1822-2311 for OVCA429. All results, * indicates p<0.05, ** indicates p<0.001, *** indicates p<1X10−10 via one-way ANOVA significance testing.

We used MMS to induce DNA damage in SKOV3 and OVCA429 cells. As expected, normalized intensity for RPA1 and RAD51 values (Fig. S5) did not display a dose dependence on MMS concentration. In SKOV3 cells treated with 1 mM MMS (Fig. 5), RPA1 recruitment to sites of DNA damage is observed alongside the diffuse signal of the nuclei. We first quantified the dose-dependent distribution of RPA1 with both Imaris Spots analysis (Fig. 5) and ICS DA measurements (Fig. 5). In SKOV3 cells, both Spots analysis (Fig. 5G) and ICS DA analysis (Fig. 5H) captured a dose dependent increase in RPA1 clustering, with the repeatability between experiments being much higher in ICS DA analysis suggesting that differences in immunofluorescence labeling between experiments impact Spots more than DA. In OVCA429 cells, Spots analysis demonstrated a more modest increase with increasing MMS dose (Fig. 5I), however, DA analysis produced a significant, larger increase in aggregation exceeding 2.5x of control at 1 mM MMS (Fig. 5J). Analysis of γH2AX clustering in OVCA429 cells treated with MMS (Fig. 4G) demonstrates a large increase at 1 mM MMS, suggesting that RPA1 DA analysis (Fig. 5J) is capturing DNA damage with a similar sensitivity to γH2AX analysis. The increased sensitivity in ICS DA versus Spots also indicates that measuring aggregation instead of foci counting could be more sensitive when analyzing proteins that do not form distinct foci.

Similar to RPA1, RAD51 does not form distinguishable foci in the nuclei in cells without pre-extraction (Fig. 5). With MMS treatment, Spots analysis of RAD51 foci captured an MMS dose dependency in OVCA429, but not SKOV3 cells (Fig. 5). However, ICS DA analysis captured dose dependent increases in aggregation (Fig. 5), with a 3X increase over control in SKOV3 at 1 mM MMS. Similar to RPA1 labeling experiments, the MMS dose dependence observed with RAD51 DA analysis is comparable to trends observed with γH2AX DA analysis for both cell lines. Thus, ICS DA proved to be more sensitive to RPA1 and RAD51 aggregation in response to DNA damage that induces significant γH2AX foci formation.

We then sought to determine if ICS analysis could provide insight into the DDR that had been previously unknown. Surprisingly, when we measured RPA1 DA induced by PARPi treatment we found different trends in different cell lines (Fig. 6). HCC1937 cells show a positive dependency of RPA1 DA with olaparib concentration. However, HT1080 cells show the opposite relationship, with RPA1 DA decreasing with increasing olaparib concentration. We then expanded our analysis to more cell lines and measured RPA1 DA, normalized to control, in response to 1 μM treatment of three different PARPi (Fig. 6B). Plotting the RPA1 DA value for each PARPi as a function of the cellular sensitivity (cell IC50) to each drug generated a curious pattern (Fig. 6C). Here, cell lines that were most sensitive to PARPi treatment demonstrated a decreasing RPA1 DA with drug potency (inverse of cell IC50), while cell lines that were more resistance had the opposite response. Fitting a linear curve to the RPA1 DA vs. cell IC50 data for each cell line and plotting the slope of that fit against the average cell line sensitivity to PARPi treatment produced a striking pattern (Fig. 6D). These results suggest the capacity to recruit RPA1 into the DDR imparts resistance to PARPi treatment. Surprisingly, the expression level of RPA1 in each cell line had no correlation to cell line sensitivity to PARPi (Fig. 6E&F). Therefore, these results demonstrate the power of ICS to identify DDR responses that would be otherwise very difficult to detect.

RPA1 recruitment correlates with cell sensitivity to PARPi.

(A) RPA1 DA as a function of olaparib concentration for HT1080 and HCC1937 cells, shown are average with SEM, n >= 504 cells, 3 biological repeats, and linear fit. (B) RPA1 DA in cells treated with 1 μM PARPi overnight, shown are average, normalized to untreated control, with SEM, n >= 350 cells, 2 biological repeats. (C) RPA1 DA as a function of cell line IC50 for each cell line and PARPi shown in Fig. 5b. Shown are average, normalized to untreated control, with SEM, n >= 350 cells, 2 biological repeats and linear fit (prism) for each cell line. (D) The linear fit slope of RPA DA vs. cell line IC50 as a function of average PARPi z-score for each cell line. Shown are slope fit with SEM and average z-score over three PARPi with st. dev. (E) Western blot of RPA1 expression levels for each cell line, with GAPDH loading control. (F) Normalized protein expression level in each cell line as a function of PARPi z score. Shown are average z-score over three PARPi with st. dev.

ICS can be Applied to a Diverse Set of Imaging and DNA Damage Response Measurements

Since ICS analysis is used on standard fluorescence microscopy images and we obtain a DAPI image every experiment we sought to incorporate cell cycle classification into measurements. Here, to demonstrate the broad applicability of ICS we performed measurements in 384 well plates on a widefield microscope. Previous results have established that DAPI intensity can be used to stage cells40. We confirmed our ability to distinguish cell stage through DAPI intensity by treating SKOV3 cells with palbociclib, camptothecin or etoposide. Similar to previous findings40, we found that palbociclib stalled cells in G1, camptothecin stalled cells in S and etoposide stalled cells in G2 (Fig. 7A). Once confirmed, we set DAPI intensity thresholds for each cell cycle stage. Thresholds were established automatically for each experiment by finding the peak of the DAPI intensity histogram of all control cells. We then classified G1 phase as cells with DAPI intensities within 10% of the DAPI intensity at the histogram peak, S phase as cells with DAPI intensity within 15% of the histogram peak times 1.5, and G2 phase as cells with DAPI intensity within 20% of the histogram peak times 2 (Fig. 7B).

Incorporating and validating cell cycle data.

(A) Histograms of DAPI intensities for cells treated with palbociclib, camptothecin or etoposide, 1 μM overnight. (B) Representative assignment of cell cycle based on DAPI histogram (grey), using the first peak as G1 cells. Normalized PCNA intensity (C) and DA (D) in SKOV3 cells that had been classified by DAPI intensity into phases of the cell cycle. Normalized γH2AX intensity (E) and DA (F) in SKOV3 cells that had been classified by DAPI intensity into phases of the cell cycle. Shown are average n>200 cells per classification and standard deviation.

We validated our cell cycle staging approach through immunofluorescence of γH2AX and PCNA. Untreated SKOV3 and HCC1937 cells were classified by cell cycle stage and both ICS DA and intensity were calculated for the two labels. Data were normalized to unclassified average. We found that PCNA intensity increased from G1 to S, which was similar to G2 for both cell lines (Fig. 7C), suggesting that PCNA expression levels increase during S phase and plateau. However, PCNA DA increased from G1 to S, then decreased from S to G2 for both cell lines (Fig. 7D), suggesting that PCNA has the highest aggregation in S phase, as would be expected. PCNA is recruited to DNA during synthesis, which will increase the overall aggregation as measured here. We found that γH2AX intensity increased from G1 to S, then plateaued in G2 for HCC1937 cells but increased from S to G2 in SKOV3 cells (Fig. 7E). γH2AX DA increased from G1 to S and from S to G2 for both cell lines (Fig. 7F). These γH2AX results confirm that γH2AX is more prominent in S and G2 than G1.

We then measured γH2AX DA response in SKOV3 cells treated with DNA damaging agents olaparib, bleomycin or etoposide (Fig. 8). Cells were classified by cell cycle and the data aggregated and plotted as a function of drug concentration. Here we observed significant changes in both γH2AX DA and intensity as a function of damaging agent concentration. However, the γH2AX DA plateaued at higher doses, while the γH2AX intensity had a consistent dose dependent increase. These measurements suggest that the amount of γH2AX in the cell is a function of the degree of DNA damage, but the clustering of γH2AX reaches a peak at very large amounts of damage. Interestingly, we did not observe large differences between cell cycle stages for any of the compounds. Overall, olaparib and etoposide generated more γH2AX intensity in S and G2 phases while bleomycin generated the highest γH2AX intensity at 10 μM in G1 cells. However, the number of cells in each cell cycle was very dependent on the drug and concentration, with all compounds inducing more cells to stall at G2.

Cell cycle analysis of DDR ICS.

Normalized γH2AX DA and intensity of SKOV3 cells treated with DMSO or olaparib (A), bleomycin (B) or etoposide (C) at 3 different doses (in μM) for 24 hours and separated by cell cycle stage based on DAPI intensity. Data were normalized to cell cycle independent DMSO average. Also shown are the ratios of the number of cells in G2 to G1 for each treatment condition. Data are n=7 biological repeats (box = average) with single cell values and standard deviation, n>200 cells per condition. For clarity all data are significantly different, p<0.05 Student’s t test from the next lower concentration unless marked – ns = no significance, dec = significant decrease.

Discussion

Fundamentally, DDR foci are resolvable when the labeled protein concentration difference between foci and the rest of the nucleus is high enough to distinguish in fluorescence microscopy. The local concentration at sites of DNA damage is driven by either protein recruitment, or, in the case of γH2AX, post translational modifications41. Yet, most DDR proteins do not cluster in DNA damage at concentrations high enough to resolve recruitment in the absence of massive, artificial DNA damage. γH2AX is the most prominent marker of DNA damage foci largely because H2AX phosphorylation occurs at sites of double stranded breaks producing a marker with very low background signal. However, recent super resolution results identify γH2AX foci as groups of smaller foci22, suggesting that not all foci are equal, and large foci in diffraction-limited microscopy represent a greater degree of DNA damage than smaller foci. Therefore, foci count alone may not accurately represent DNA damage.

Clustering analysis with ICS overcomes both limits of counting traditional markers and measuring non-traditional markers as metrics of DNA damage or to study their role in the DDR. Initially we used the TNoP parameter to match foci counts in individual cells from two methods, however ICS calculated degree of aggregation (DA) captures overall clustering of protein and is therefore the more appropriate ICS measurement. Using ICS DA, we were able to detect the dose dependent response to two PARP inhibitors, matching or exceeding dose-dependent sensitivity of foci counting via Imaris Spots. Furthermore, ICS DA was uniquely able to measure dose-dependent DNA damage using RPA1 and RAD51 as DNA damage markers. The accuracy of these measurements was validated by the similarity of the dose dependency to γH2AX measurements. Therefore, ICS is able to evaluate DNA damage with non-traditional markers or evaluate the recruitment of proteins to DNA damage.

ICS analysis of γH2AX in OVCA429 cells treated with PARP inhibitors, where there was not a dose dependent increase of γH2AX intensity, revealed an increase in DA, indicating larger foci cluster formation was a merging of smaller clusters to produce an overall decrease in the number of foci clusters35,36. The formation of large foci can impact foci counting methods, however, Imaris Spots uses a spot size parameter that aids in placing multiple smaller spots per large foci. Yet, this spot size parameter failed to consistently capture the dose dependent increases in DNA damage where γH2AX intensity was not also increased. Spatial ICS is well suited to measure large aggregations and has previously been used to count distinguishable objects larger than the diffraction limit such as fluorescent beads28 and dendritic spines42, by using an intensity threshold to remove the background signal. This can be supplemented by various image analysis algorithms such as Gaussian filters to smooth out fluctuations in the objects so they can be detected as one object. However, we did not need an intensity threshold to remove the background signal, and only used a background subtraction that represented the intensity in a part of the image with no cells to account for autofluorescence outside of the cells.

We then used ICS to assess the involvement of RPA1 in the DDR, with results suggesting that recruitment correlates with cellular sensitivity to PARPi. These results were not apparent when measure RPA1 expression levels across cell lines, a typical evaluation of cell line sensitivity, indicating that ICS is capable of revealing underlying biology not detected by traditional approaches. We also demonstrated that cell cycle analysis can be readily performed on results to classify cells by cell stage for measurements of the impact of cellular stage on the DNA damage response. As a simple image analysis technique, ICS can be applied broadly to incorporate other analysis as well.

Here, we demonstrated that ICS is an alternative technique to foci counting, where the clustering involved in protein recruitment during DNA damage can be fully captured on the molecular level. ICS measures the degree of clustering and considers the low intensity signal contributions that are often lost during foci segmentation. Our findings correlate with prior studies characterizing foci formation/ clustering and the DNA damage response using advanced optical techniques such laser micro-irradiation43 and super resolution microscopy22. Advantageously, ICS can be implemented on any fluorescence image as long as the square relative fluorescence intensity fluctuations are detectable above noise fluctuations. ICS does not require laser micro-irradiation to induce detectable clustering23,43,44. Also, it overcomes the complications introduced when using pre-extraction in immunofluorescence45, where soluble protein is removed through permeabilizing the cell for an arbitrary amount of time, likely removing some recruited protein as well. Thus, ICS enables simple, standard immunofluorescence labeling techniques. The overall approach, using a combination of conventional fluorescence microscopy, antibody labeling, and ICS analysis, provides a molecular level understanding for characterizing protein recruitment and signaling in a variety of applications from capturing the DNA damage response to evaluating cancer therapies. We expect adoption of this approach will both lead to a more objective measure of DNA damage and provide a tool to evaluate the role of every DDR protein during the DNA damage response.

eLife Assessment

This valuable paper shows image correlation spectroscopy (ICS) as a new tool to analyze the clustering of proteins involved in DNA damage response (DDR). The solid evidence presented demonstrates that this method is more sensitive than traditional focus counting, although some of the claims require further contextualization. This new method provides an alternative tool to analyze immuno-stained focus for researchers in the fields of DDR and cell biology.

Reviewer #1 (Public review):

Summary:

This manuscript assesses the utility of spatial image correlation spectroscopy (ICS) for measuring physiological responses to DNA damage. ICS is a long-established (~1993) method similar to fluorescence correlation spectroscopy, for deriving information about the fluorophore density that underlies the intensity distributions of images. The authors first provide a technical but fairly accessible background to the theory of ICS, then compare it with traditional spot-counting methods for its ability to analyze the characteristics of γH2AX staining. Based on the degree of aggregation (DA) value, the authors then survey other markers of DNA damage and uncover some novel findings, such as that RPA aggregation inversely tracks the sensitivity to PARP inhibitors of different cell lines.

The need for a more objective and standardized tool for analyzing DNA damage has long been felt in the field and the authors argue convincingly for this. The data in the manuscript are in general well-supported and of high quality, and show promise of being a robust alternative to traditional focus counting. However, there are a number of areas where I would suggest further controls and explanations to strengthen the authors' case for the robustness of their ICS method.

Strengths:

The spatial ICS method the authors describe and demonstrate is easy to perform and applicable to a wide variety of images. The DDR was well-chosen as an arena to showcase its utility due to its well-characterized dose-responsiveness and known variability between cell types. Their method should be readily useable by any cell biologist wanting to assess the degree of aggregation of fluorescent tags of interest.

Weaknesses:

The spatial ICS method, though of longstanding history, is not as intuitive or well-known as spot-based quantitation. While the Theory section gives a standard mathematical introduction, it is not as accessible as it could be. Additionally, the values of TNoP and DA shown in the Results are not discussed sufficiently with regard to their physical and physiological interpretation.

The correlation of TNoP with γH2AX foci is high (Figure 2) and suggestive that the ICS method is suitable for measuring the strength of the DDR. The authors correctly mention that the number of spots found using traditional means can vary based on the parameters used for spot detection. They contrast this with their ICS detection method; however, the actual robustness of spatial ICS is not given equal consideration.

Reviewer #2 (Public review):

Summary:

Immunostaining of chromatin-associated proteins and visualization of these factors through fluorescence microscopy is a powerful technique to study molecular processes such as DNA damage and repair, their timing, and their genetic dependencies. Nonetheless, it is well-established that this methodology (sometimes called "foci-ology") is subject to biases introduced during sample preparation, immunostaining, foci visualization, and scoring. This manuscript addresses several of the shortcomings associated with immunostaining by using image correlation spectroscopy (ICS) to quantify the recruitment of several DNA damage response-associated proteins following various types of DNA damage.

The study compares automated foci counting and fluorescence intensity to image correlation spectroscopy degree of aggregation study the recruitment of DNA repair proteins to chromatin following DNA damage. After validating image correlation spectroscopy as a reliable method to visualize the recruitment of γH2AX to chromatin following DNA damage in two separate cell lines, the study demonstrates that this new method can also be used to quantify RPA1 and Rad51 recruitment to chromatin following DNA damage. The study further shows that RPA1 signal as measured by this method correlates with cell sensitivity to Olaparib, a widely-used PARP inhibitor.

Strengths:

Multiple proof-of-concept experiments demonstrate that using image correlation spectroscopy degree of aggregation is typically more sensitive than foci counting or foci intensity as a measure of recruitment of a protein of interest to a site of DNA damage. The sensitivity of the SKOV3 and OVCA429 cell lines to MMS and the PARP inhibitors Olaparib and Veliparib as measured by cell viability in response to increasing amounts of each compound is a valuable correlate to the image correlation spectroscopy degree of aggregation measurements.

Weaknesses:

The subjectivity of foci counting has been well-recognized in the DNA repair field, and thus foci counts are usually interpreted relative to a set of technical and biological controls and across a meaningful time period. As such:

(1) A more detailed description of the numerous prior studies examining the immunostaining of proteins such as γH2AX, RAD51, and RPA is needed to give context to the findings presented herein.

(2) The benefits of adopting image correlation spectroscopy should be discussed in comparison to other methods, such as super-resolution microscopy, which may also offer enhanced sensitivity over traditional microscopy.

(3) Additional controls demonstrating the specificity of their antibodies to detection of the proteins of interest should be added, or the appropriate citations validating these antibodies included.

Reviewer #3 (Public review):

Summary:

This paper described a new tool called "Image Correlation Spectroscopy; ICS) to detect clustering fluorescence signals such as foci in the nucleus (or any other cellular structures). The authors compared ICS DA (degree of aggregation) data with Imaris Spots data (and ImageJ Find Maxima data) and found a comparable result between the two analyses and that the ICS sometimes produced a better quantification than the Imaris. Moreover, the authors extended the application of ICS to detect cell-cycle stages by analyzing the DAPI image of cells. This is a useful tool without the subjective bias of researchers and provides novel quantitative values in cell biology.

Strengths:

The authors developed a new tool to detect and quantify the aggregates of immuno-fluorescent signals, which is a center of modern cell biology, such as the fields of DNA damage responses (DDR), including DNA repair. This new method could detect the "invisible" signal in cells without pre-extraction, which could prevent the effect of extracted materials on the pre-assembled ensembles, a target for the detection. This would be an alternative method for the quantification of fluorescent signals relative to conventional methods.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript assesses the utility of spatial image correlation spectroscopy (ICS) for measuring physiological responses to DNA damage. ICS is a long-established (~1993) method similar to fluorescence correlation spectroscopy, for deriving information about the fluorophore density that underlies the intensity distributions of images. The authors first provide a technical but fairly accessible background to the theory of ICS, then compare it with traditional spot-counting methods for its ability to analyze the characteristics of γH2AX staining. Based on the degree of aggregation (DA) value, the authors then survey other markers of DNA damage and uncover some novel findings, such as that RPA aggregation inversely tracks the sensitivity to PARP inhibitors of different cell lines.

The need for a more objective and standardized tool for analyzing DNA damage has long been felt in the field and the authors argue convincingly for this. The data in the manuscript are in general well-supported and of high quality, and show promise of being a robust alternative to traditional focus counting. However, there are a number of areas where I would suggest further controls and explanations to strengthen the authors' case for the robustness of their ICS method.

Strengths:

The spatial ICS method the authors describe and demonstrate is easy to perform and applicable to a wide variety of images. The DDR was well-chosen as an arena to showcase its utility due to its well-characterized dose-responsiveness and known variability between cell types. Their method should be readily useable by any cell biologist wanting to assess the degree of aggregation of fluorescent tags of interest.

Weaknesses:

The spatial ICS method, though of longstanding history, is not as intuitive or well-known as spot-based quantitation. While the Theory section gives a standard mathematical introduction, it is not as accessible as it could be. Additionally, the values of TNoP and DA shown in the Results are not discussed sufficiently with regard to their physical and physiological interpretation.

We agree that a major limitation in adaption of this approach is a deeper understanding of the theory and results. We have updated the theory section to include further discussion (Page 4 line 132)

The correlation of TNoP with γH2AX foci is high (Figure 2) and suggestive that the ICS method is suitable for measuring the strength of the DDR. The authors correctly mention that the number of spots found using traditional means can vary based on the parameters used for spot detection. They contrast this with their ICS detection method; however, the actual robustness of spatial ICS is not given equal consideration.

We found it difficult to give equal consideration of robustness to ICS. The major limitation of traditional approaches is proper selection of an intensity threshold that is necessary to define and separate foci from background intensity. However, ICS does not employ a threshold, therefore we could not test different thresholding applications in ICS as we did with traditional methods. In our view the absence of the need for a threshold is profoundly advantageous. The only inputs we employ in the ICS analysis are used to segment cell nuclei, yet these have no impact on the ICS calculation and are necessary for any analysis of the DDR.

Reviewer #2 (Public review):

Summary:

Immunostaining of chromatin-associated proteins and visualization of these factors through fluorescence microscopy is a powerful technique to study molecular processes such as DNA damage and repair, their timing, and their genetic dependencies. Nonetheless, it is well-established that this methodology (sometimes called "foci-ology") is subject to biases introduced during sample preparation, immunostaining, foci visualization, and scoring. This manuscript addresses several of the shortcomings associated with immunostaining by using image correlation spectroscopy (ICS) to quantify the recruitment of several DNA damage response-associated proteins following various types of DNA damage.

The study compares automated foci counting and fluorescence intensity to image correlation spectroscopy degree of aggregation study the recruitment of DNA repair proteins to chromatin following DNA damage. After validating image correlation spectroscopy as a reliable method to visualize the recruitment of γH2AX to chromatin following DNA damage in two separate cell lines, the study demonstrates that this new method can also be used to quantify RPA1 and Rad51 recruitment to chromatin following DNA damage. The study further shows that RPA1 signal as measured by this method correlates with cell sensitivity to Olaparib, a widely-used PARP inhibitor.

Strengths:

Multiple proof-of-concept experiments demonstrate that using image correlation spectroscopy degree of aggregation is typically more sensitive than foci counting or foci intensity as a measure of recruitment of a protein of interest to a site of DNA damage. The sensitivity of the SKOV3 and OVCA429 cell lines to MMS and the PARP inhibitors Olaparib and Veliparib as measured by cell viability in response to increasing amounts of each compound is a valuable correlate to the image correlation spectroscopy degree of aggregation measurements.

Weaknesses:

The subjectivity of foci counting has been well-recognized in the DNA repair field, and thus foci counts are usually interpreted relative to a set of technical and biological controls and across a meaningful time period. As such:

(1) A more detailed description of the numerous prior studies examining the immunostaining of proteins such as γH2AX, RAD51, and RPA is needed to give context to the findings presented herein.

We apologize for not providing enough detail. We have added further references and discussion. γH2AX foci counting, in particular, has been used in thousands of previous studies. (Pages 18 line 513 and 517)

(2) The benefits of adopting image correlation spectroscopy should be discussed in comparison to other methods, such as super-resolution microscopy, which may also offer enhanced sensitivity over traditional microscopy.

Thank you for raising this point. We have added this discussion (page 19 line 553). The limiting factor that ICS addresses is the partition coefficient of signal in a foci or cluster versus outside the cluster. Super-resolution will not necessarily improve this unless it is resolved down to single molecule counting. However, one would still need to evaluate how to define a cluster or foci in the background of non-cluster distribution.

(3) Additional controls demonstrating the specificity of their antibodies to detection of the proteins of interest should be added, or the appropriate citations validating these antibodies included.

We have added text stating that we only use validated antibodies (page 6 line 193). One thing to note is that we are measuring differences between treatment conditions, thus, if an antibody has non-specific labeling of proteins of cellular structures that do not change upon treatment, our approach would overcome this limitation.

Reviewer #3 (Public review):

Summary:

This paper described a new tool called "Image Correlation Spectroscopy; ICS) to detect clustering fluorescence signals such as foci in the nucleus (or any other cellular structures). The authors compared ICS DA (degree of aggregation) data with Imaris Spots data (and ImageJ Find Maxima data) and found a comparable result between the two analyses and that the ICS sometimes produced a better quantification than the Imaris. Moreover, the authors extended the application of ICS to detect cell-cycle stages by analyzing the DAPI image of cells. This is a useful tool without the subjective bias of researchers and provides novel quantitative values in cell biology.

Strengths:

The authors developed a new tool to detect and quantify the aggregates of immunofluorescent signals, which is a center of modern cell biology, such as the fields of DNA damage responses (DDR), including DNA repair. This new method could detect the "invisible" signal in cells without pre-extraction, which could prevent the effect of extracted materials on the pre-assembled ensembles, a target for the detection. This would be an alternative method for the quantification of fluorescent signals relative to conventional methods.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Major comments:

(1) The ICS theory section is essential and based on an excellent review from one of the authors. It would benefit greatly from a diagram showing where the quantities 𝒈(𝟎, 𝟎), 𝝎𝟎, and 𝒈inf come from in the 2D Gaussian fit, ideally for two cases where these quantities differ (i.e., how they correspond to different DA or TNoP values). In my opinion, this addition would greatly increase the manuscript's accessibility for DDR researchers. The citation of the review at the beginning would also be a plus.

We have added the review citation at the front of the theory section (page 3 line 87).We have highlighted where g(0,0), the most critical measurement for determination of TNoP and DA, derives from in Figure 2D. However, it is difficult to describe all the curve fit parameters in an image as they have some interdependency on each other and thus labeling one in a single image would not independently capture how they might be observed in a different curve fit.

(2) The TNoP measured in Figure 2 is a quantity about 2000-3000 times greater than the number of "traditionally detected" foci by both methods and the linear relations have very low Y intercepts. Can the authors comment explicitly on the physical interpretation of this number - are 2 to 3 thousand independent particles present within each "focus" detected by traditional means? If so, then what might one "particle" correspond to? (a single secondary antibody or fluorophore? a nucleosome?). In a similar vein, the X intercepts lie at around 25 foci, meaning that in images with fewer than that number of foci detected by ImageJ or Imaris, the ICS method should detect zero TNoP - is this in line with the authors' predictions? Is it possible that a first-order line fit is not the most appropriate relation between the two methods?

We apologize for our brevity here. Since DA proved to be a more useful metric we did not spend much effort discussing TNoP. TNoP correlates to the number of clustered particles, or non-diffuse fluorophores. TNoP is the inverse of the number of individual particles per nucleus, but the value is not a direct measure of foci. If a sample had no clustering at all, the number of individual particles would be at a maximum and the TNoP would be at a minimum. However, as fluorophores cluster, the number of individual particles (i.e. non-clustered fluorophores) decreases, which increases the TNoP value. Therefore, TNoP has a correlation to the number of foci detected through traditional measurements, as we found here. Yet, TNoP is a relative measurement and cannot be compared across different conditions. Similar to foci counting, TNoP is unable to factor the size or intensity of each cluster, thus DA is a more appropriate quantification of the DNA damage response.

The value of TNoP is dependent on the fitted point spread function and the area of the nucleus. The y=0 intercept of TNoP is defined by the optical setup and is not expected to necessarily go through x=0. Intriguingly, other groups have found that some foci identified through traditional measurements are actually clusters of multiple smaller foci, thus the concept of what a foci represents is difficult to interpret. Thus, here we aimed to show a general correlation of TNoP with foci count through traditional methods to reflect how ICS is similar to foci counting, then employed DA to overcome the limitations of defining a foci.

We have tried to clarify this in the text (page 8, line 266)

(3) Some suggestions to address the robustness of ICS:

For a given sample (i.e. one segmented nucleus), the calculation of DA and TNoP should be similar between different images of that same nucleus taken at different times, similar to how the number of traditionally detected foci would be fairly invariant. In particular, it should be shown that these values are not just scaling with the higher normalized intensity seen in stronger DDR responses. In the same vein, the linear relationship between TNoP and "foci" should not change even if the confocal settings are slightly different (i.e., higher/lower illumination intensity) as long as the condition stipulated by the authors in the Discussion holds ("ICS can be implemented on any fluorescence image as long as the square relative fluorescence intensity fluctuations are detectable above noise fluctuations."). To show, as the title states, that spatial ICS is a robust tool, it would be desirable to demonstrate this with a series of images of the same cell at the same or varying excitation intensities.

Thank you for your suggestions. Indeed, the calculation will be the same over sequential images of the same cell. Observations of dose dependent DA that does not correlate with intensity for RPA1 and RAD51 results (Fig. S5) directly demonstrates that DA does not just scale with intensity.

We would not expect the TNoP to change with confocal setting, however we show in Figure 1 that the number of foci does indeed change with intensity settings as captured by thresholds. Therefore, any interpretation of TNoP vs. foci count would be very difficult to make at different microscope settings. To ensure we are fairly comparing ICS to existing analysis we keep the settings the same and measure changes between conditions.

(4) More information is needed on how intensity normalization was performed. The Methods states "Measurements across experiments were normalized by the control in each dataset." The DMSO (0mM drug) plots all appear to have a mean of 1.0, so it appears the values for each set of control nuclei were divided by their own mean, and then the values for each set of experimental nuclei were divided by the mean value of all 3 controls as an aggregate; is this correct?

We apologize for not being more clear. Thank you for raising this point. We normalized data to a control from each experimental group. Thus, in figures 3,4 and 5 data were collected over multiple experiments with one control per experiment and each treatment condition included in each experiment. Therefore, we normalized each result to the corresponding control from that imaging session. However, in Figure 8 we ran experiments at much higher throughput with multiple controls per experiment, thus the data were normalized to the overall average of the controls, which is why the control averages are not all at a value of 1. We have clarified this in the text. (Page 7 line 218).

(5) Some more information about the ICS analysis should be given if the full code is not provided - in particular, how the nucleus mask was implemented on the "signal" channel (were the edges abruptly set to zero or was a window function introduced to avoid edge effects in the discrete FFT?

Thank you for raising this point. We have added the code to GitHub - github.com/ dubachLab/ics. The signal region was established by simply applying the nuclear mask from the DAPI channel to the IF channel. Each region is padded with average intensity value at the edges for 2x the dimensions of the ROI to remove edge effects in the FFT.

Minor comments:

(1) Figure 3, 4, 5: I think it would aid figure readability if channels were labeled in the images themselves, not just in the legend.

Thank you for the suggestion, we tried doing this and struggle to fit a label with the layout of the images. We were also concerned about interpretation of data in each column and the potential to assign data to each figure if they were so prominently labeled.

(2) Supplemental Figures are mislabeled; the order given in the legends is S1, S2, S3, S2, S3. S4 is called out in the main text where it should be S5.

Thank you for catching this error. We have made the necessary corrections. S4 contains data on cellular response to the drugs, while S5 contains intensity data in response to MMS.

(3) It should be stated for each Figure what kind of microscopy was performed - I assume that it is confocal for everything except when widefield is explicitly stated, but for clarity please add this information.

Indeed, this is correct, we have indicated which microscopy was used for each figure.

(4) The MATLAB code and full (uncropped) Western blots should be provided as supplemental data if possible.

We have included a GitHub link for the code and un-cropped western blots.

(5) The p values from significance tests should indicate whether multiple comparisons correction was necessary (if suggested by Prism) and performed.

Apologies for a lack of clarity but this was not necessary, significance was calculated vs. the next lower dose (e.g. 10 micromolar vs. 1 micromolar). We have clarified this in the methods (page 7 line 221).

Reviewer #2 (Recommendations for the authors):

Major points:

In addition to the weaknesses noted above, to encourage widespread adoption of this method, the authors should make the tools that they used for their analysis publicly available. In a few instances (e.g., compare Figures 3J and 3L), other methods outperform DA. It would be meaningful to discuss when especially DA may be a better measure than others (such as intensity or number of foci).

We have made code available on Github. We expect results, such as those in Figures 3J and 3L where intensity is significantly higher at the highest concentration but DA is not are reflective of the underlying biology and this may be interpreted differently under different experimental conditions. Imaris spots (Fig. 3K) also does not capture a significant increase at the highest dose of olaparib, suggesting that intensity may raise but it doesn’t not generate more foci. These results are likely highly dependent on the mechanism of olaparib at such a high concentration and the DDR response. We are hesitant to draw biological conclusions from these results and instead would like to highlight the capacity of ICS to evaluate the DDR, therefore we don’t want to make any broad comments about different applications.

Minor points:

(1) Pg. 12: "We used MMS to induce DNA damage in SKOV3 and OVCA429 cells. As expected, normalized intensity for RPA1 and RAD51 values (Figure S5) did not display a dose dependence on MMS concentration."

Please provide a citation for the claim that RPA1 and RAD51 normalized intensities do not display a dose dependence on MMS concentration.

These were data that we generated. We were not expecting an intensity change as that would presumably require increased protein generation in response to MMS, compared to gH2AX where the phospho-specific H2AX is generated in the DDR.

(2) Pg. 12: "Similar to RPA1, RAD51 does not form distinguishable foci in the nuclei in cells without preextraction (Fig. 5)." Please provide a citation for this claim.

We did not do pre-extraction and our results don’t produce changes in distinguishable foci. We provided citations discussing how, without pre extraction, foci formation for these proteins is not obvious (REF 38 and 39).

(3) I noted that the authors cite one paper [38] apparently showing that RPA and Rad51 do not always form foci, however, this is in the C. elegans germline in response to micro irradiation, therefore I am not sure that it is applicable to human cells.

We apologize for referencing a paper on C elegans. Most papers looking at RPA and RAD51 in the DDR use pre-extraction as it seems necessary to observe foci. Therefore, there are not as many papers, that we could find, that do not use pre-extraction. Reference 39 is in Hela cells.

Reviewer #3 (Recommendations for the authors):

Major points:

(1) Page 8, the second paragraph: In the Result section, it is better to describe how the authors carried out immuno-staining (without pre-extract subtraction) and ICS briefly, although the method is described in detail in the Method section.

Thank you for the suggestion, we have added this description (page 8, line 259)

(2) In Figure 5K-P: The authors analyzed "invisible" RAD51 foci on the image (Fig. 5L, M, O, and P) without pre-extraction. As a control experiment, it is useful to check whether pre-extraction would provide "visible" RAD51 foci and to examine the similar MMS concentration dependency shown in Figure 5R (or 5T). This would strengthen the power of the ICS analysis.

Thank you for the suggestion. In our hands, pre-extraction is extremely subjective. We have tried performing pre-extraction but find highly variable results depending on conditions. Therefore, we did not include any pre-extraction here. We expect that performing these experiments may or may not agree with results in Figure 5 largely because we are unable to achieve repeatable pre-extraction foci counting.

(3) Figure 6D (and 6C) looks very interesting. It would be important to show the interpretation of this correlation shown in the graph. Although the authors argued that ICS analysis results shown in the graph could provide new insight into the DDR (page 14, last line 5), as shown in another part, it is important to carry out the same analysis by using Imaris Spots. Moreover, it is interesting to apply the analysis to RAD51 foci (shown in Figure 5), given that the PARPi effect is enhanced in the absence of RAD51mediated recombination.

We completely agree that this analysis may generate interesting results to help interpret the DDR response to PARP inhibition. These experiments are part of an ongoing follow up study where we extend the use of ICS to other parts of the DDR and investigate protein clustering across several proteins with impact on PARPi response. Therefore, since the focus of this manuscript is introducing ICS as a tool to study the DDR, we believe that omitting those data here does not deter from the central points of the manuscript. We including results in Figure 6 because we wanted to show how ICS could impact DDR research. Furthermore, combined with our advances shown in Figures 7 and 8, we are currently working on adapting ICS to be high-throughput and much simpler than Imaris spots for handling large datasets needed to generate results like those in Figure 6.

Minor points:

(1) Figure 1I, blue arrows: These showed an area with a higher background. Because of a low magnification, it is very hard to see the difference from the other areas of the background. It is better to show a magnified image of the representative region with a higher background.

We hope that readers can see the higher intensity in the diffuse area. We attempted to construct a zoomed in area, but that either blocked a significant portion of the nonzoomed image or added complexity to the figure. We have noted that images in Figure S1 are larger and more obviously capture an increase in background intensity.

(2) Figure 2 legend, line 5, the same as "A)": This should be "B".

Here, the number of independent particle clusters is intended to be the same as A, the difference is that the independent particles are clusters in C and individual fluorophores in A.

(3) Page 9, the first paragraph, last line, foci formation, and foci composition: These should be "focus formation and focus composition".

We have changed this.

(4) Page 15, the first paragraph, line 5, palbociclib, camptothecin, or etoposide: please explain what kinds of the drugs are.

We have added that these drugs cause cells to stall at different cell cycle stages. Explaining the drugs would take considerable room in the text.

(5) Page 16, the first paragraph, line 1, bleomycin: Please explain what this drug is.

Similar to above, we have stated that this drug causes DNA damage, going into detail would take several sentences.

Version 2

Abstract

Abstract

The DNA Damage response (DDR) is both essential and highly complex. Evaluating the DDR is a critical aspect of cell biology. Counting DNA damage foci is one of the most common approaches to study the DDR. Yet, quantification of protein foci suffers from experimental limitations, subjectivity of analysis and is restricted to a handful of the hundreds of DDR proteins. Here we apply image correlation spectroscopy (ICS) to quantify the local clustering at sites of DNA damage directly. We found that ICS outperformed foci counting of traditional DDR markers and enabled quantification of other markers without the complex labeling procedures that are otherwise required. ICS analysis also provided insight into DDR protein recruitment that was previously undetectable. Further expansion incorporating analysis to cell cycle classification demonstrates a rapid, non-biased approach to fully study the DNA damage response within cells. ICS analysis presents an objective, quantitative image analysis technique to study the DNA damage response in unaltered cells that we expect will significantly enhance quantitative DNA damage response research.

Introduction

The DNA damage response (DDR) represents multiple different pathways that are dependent on several factors including cell cycle and the type of damage1-3. DNA damage and the DDR are prominent in several disease areas such as cancer4, neurodegeneration5 and aging6, and the aberrant DDR is a major target of cancer therapies7. Collectively, the DDR comprises hundreds of proteins that are recruited to sites of damage in a concerted manner serving as scaffolds, DNA binders, modifying enzymes and repair enzymes. Largely due to the massive evolutionary pressure to maintain the genome, the DDR is enormously complex, rendering it difficult to study in cells.

Current methods to evaluate DNA damage vary from molecular strategies such as polymerase chain reaction8, comet assays9 and gel electrophoresis10 to intracellular approaches that include, fluorescence in situ hybridization11, radioimmunoassay12 and TUNEL assays13. Analytical methods have also been employed, for example, gas or liquid chromatography, mass spectrometry and electrochemical approaches have been used to identify DDR components and signaling7. However, the gold standard for assessing DNA damage levels in cells is fluorescence imaging and counting DNA repair foci formation by labeling established protein markers, such as Ku70/8014, γH2AX15 or XRCC116, through immunofluorescence or genetically encoded fluorescent proteins.

Historically, foci counting was performed manually and relied on individual researchers to identify foci, introducing an inherent bias to the process. More recently, foci counting has been widely automated via computer-based algorithms17,18. Numerous software tools have been released to count foci via methods of signal thresholding, maxima finding, and morphological reconstructions to accurately detect foci in nuclei. However, foci are not often individually distinguishable and are present at varying sizes and intensities, leading to some ambiguity in thresholding, segmenting and counting parameters, with a wide variety of software employing different algorithms19-21. Furthermore, γH2AX foci were recently revealed to represent clustered structures of γH2AX nano-domains22, suggesting multiple sites of DNA damage are present in single foci. Thus, it is likely that not all foci are the same – larger foci representing a greater degree of or more mature DNA damage than smaller foci.

In addition to measurements of known foci markers, studies using laser micro-irradiation have shown that hundreds of different proteins are recruited to sites of DNA damage23, despite not forming foci that are resolvable in fluorescence images when more biologically relevant DNA damage is induced. Therefore, the role of the majority of DDR proteins remains poorly understood because their recruitment to DNA damage foci is difficult to quantify with traditional approaches. Typically, non-recruited protein is pre-extracted, through lysis of the cell and nucleus, prior to fixation to reduce the background concentration and enhance the resolution of protein foci. However, the demonstration that DNA damage foci are biomolecular condensates where foci protein concentration exists in equilibrium with the nuclear concentration24 suggests that pre-extraction may impact the DNA damage foci as well to hinder the accuracy of the results.

To overcome the limitations of current cellular DDR analysis approaches, we hypothesized that measuring clustering of labeled proteins within the nucleus could quantify the degree of DNA damage and involvement of any DDR protein. To test this approach, we employed fluorescence microscopy imaging with fluorescence fluctuation analysis via spatial image correlation spectroscopy (ICS) as a tool for measuring the clustering of proteins during DDR in multiple cancer cell lines treated with DNA damaging agents and DDR therapeutics. Fluorescence correlation spectroscopy (FCS), which requires temporal imaging, has been previously used for the measurements of macromolecular aggregation via autocorrelation analysis of fluorescence intensity fluctuations that arise due to changes in the local concentration of fluorescence molecules in a tiny focal volume25-27. Thus, FCS is well suited to study single, well-defined foci. Conversely, ICS was developed as the imaging analog of FCS to measure densities and aggregation states of membrane receptors from spatial correlation analysis of very small fluorescence fluctuations in single images28,29. Thus, ICS presents a potential alternative method to quantify DNA damage and expand the detection of DDR response elements beyond established protein markers.

Methods

Theory

Image correlation spectroscopy is based on the analysis of fluorescence intensity fluctuations arising from variations in the number of fluorescent particles within focal spots imaged in space using a fluorescence microscope30. Under the assumptions of particle ideality and linear fluorescence emission (i.e., no signal saturation or energy transfer), the detected mean fluorescence intensity of tagged molecules varies linearly with the concentration of fluorophores in a focal spot (Fig. 2A-B). The square relative intensity fluctuation (intensity variance/mean intensity, Eq. 1) will be the mean number of detected independent fluorescent particles per focal spot, since ideal behavior entails that the molecules obey Poisson statistics within the volume:

Common approaches to count DNA damage foci.

(A) DAPI channel with 5 µm scale bar, (B) γH2AX channel, (C) Composite of A) and B). (D-E) Foci counting with Find Maxima tool in ImageJ with a threshold of D) 600 and E) 300. (F) Foci counting using Imaris Spots with an automatic threshold and an expected spot size of 1 µm. (G) Number of maxima per nucleus from different thresholds using ImageJ for the same cell. (H) Foci counting comparison between a user defined threshold in ImageJ and automatic threshold in Imaris on the same cell. (I) Intensity adjusted γH2AX channel showing lower intensity foci (white arrows), larger foci (green arrows) and higher background intensities (blue arrows). Figure S1 provides larger images for further clarity.

Counting foci using image correlation spectroscopy correlates with optimized measurements using existing approaches.

(A) Schematic population of monomeric fluorescent particles where emitting particles (green) are in a diffraction limited excitation focus (cyan) – not to scale. (B) Schematic showing increase in particle density from A) which results in an increase in average fluorescence intensity. (C) Schematic showing clustering of fluorescent particles where the total number of independent particle clusters is the same as A). Spatial ICS measures the average number of independent fluorescent units per area which, along with the average fluorescence intensity, is used to calculate the average degree of aggregation of the particles. (D) Intensity-based segmentation of the DAPI channel with an automatic threshold using the MATLAB function imbinarize to create a mask of the nucelus. The mask specifies the ROI for ICS analysis of the γH2AX channel, outputting mean number of independent fluorescent particles per focal spot area (inverse of the intensity normalized spatial correlation function amplitude). (E) and (F) Comparison of the ICS parameter TNoP with E) ImageJ Find Maxima tool with a user defined threshold and F) Imaris Spots with an automatic threshold for the same nuclei. A linear line of best fit is shown on each graph with the resulting R2.

given intensity fluctuations defined as δi = i − ⟨i⟩, where ⟨i⟩ is the mean intensity and ⟨n⟩ is the mean number of independent fluorescent particles per focal spot area.

To obtain this number density information independent of white noise sources (such as shot noise), ICS calculates the mean intensity normalized spatial autocorrelation function of fluorescence intensity fluctuations (Eq. 2) of a region of interest (ROI). The orthogonal spatial lag variables ξ and η represent discrete pixel shifts in x and y directions in an image at which the spatial correlation is calculated. The zero spatial lags value of the spatial autocorrelation function is the square relative intensity and hence the particle number density.

For computational speed, the spatial correlation function is calculated using Fourier methods (Eq. 3)28, where F is the discrete 2D spatial fast Fourier transform of the ROI, F* is the complex conjugate and F-1 is the inverse Fourier transform.

The calculated spatial intensity fluctuation correlation functions are then each fit to a 2D Gaussian (Eq. 4) using a non-linear least-squares algorithm, where the zero lags point is not weighted due to white noise contributions. Output fit parameters are g(0, 0), the zero-lags amplitude, ω0, the e-2 Gaussian correlation radius, and g∞, the long spatial lag offset.

The best fit zero lags amplitude of the correlation function is an estimate of the square relative fluctuation from Eq. 1, with an inverse that is the mean number of independent fluorescent particles per focal spot area, ⟨n⟩ (Eq. 5). The total number of particles (TNoP) in the ROI can be calculated using the image area and ω0 (Eq. 6).

When fluorescent molecules undergo aggregation or clustering, the clusters/aggregates are not resolvable using diffraction limited optical microscopy (i.e., confocal) unless large aggregates are present (such as individually distinguishable foci). However, aggregation manifests in the ROI as larger relative intensity fluctuations from a smaller number of brighter fluorescent particles per focal spot. If the number of overall fluorophores does not change, the mean fluorescent intensity should be constant, but the number of independent fluorescent particles (formed of one or more fluorophores) should decrease after clustering (Fig. 2 B-C). Since the average intensity of the ROI is proportional to the total number of fluorophores ⟨nf⟩ after background correction and a constant (c) relating intensity to fluorophore count, a degree of aggregation (DA) measurement can be calculated using the average intensity, ⟨i⟩, and the mean number of fluorescent particles, ⟨n⟩;

To first order, the degree of aggregation is proportional to the mean number of fluorophores per aggregate if the variance of the aggregate distribution is small29.

The output from ICS calculations can be used to understand the distribution of fluorophores contributing to the image. The two outputs used here are TNoP and DA. TNoP is a measure of the inverse of the number of independent particles – this calculation is not dependent on the size (or number of fluorophores) contributing to each independent particle, nor the distribution of particle sizes. The DA captures the size of each independent cluster through dividing the average voxel intensity by the average number of independent particles per voxel. Thus, the DA captures how many fluorophores comprise each independent particle in the image.

Image Analysis

For the ICS analysis, nuclei were segmented from the DAPI channel using a watershed approach in MATLAB. Briefly, we applied a difference of gaussians for size-based feature extraction of nuclei in the image (upper gaussian σ=100, lower gaussian σ=2-5 dependent on cell size). This feature enhanced image was then binarized to create a primary mask (using the function imbinarize) and connected components under 100 pixels were removed (using the function bwareaopen). We then created a distance transform of the resulting binary image (using the function bwdist). Seed points were determined by finding the regional maximum of a gaussian filtered image, with a gaussian σ of appropriate size for seed point detection (σ=7 or 8 dependent on cell size). The seed points were applied to the distance transform, and we applied the watershed function to the resulting image to obtain masks for each nucleus (extracted using the function regionprops).

Using the nucleus masks, we performed ICS analysis on the antibody-stained channel, within size limitations for a nucleus (objects > 400 and < 2400 pixels). Intensity, DA and TNoP were calculated for each cell. In this analysis, since the ROI is defined by the nucleus mask (segmented from nuclear signal using DAPI staining), TNoP represents the total number of particles per nucleus, with the image area being the number of pixels in each nuclear mask.

Measurements across experiments were normalized by the control in each dataset. The confocal imaging settings were kept the same across each dataset to provide comparable measurements per dataset. Data were plotted in Prism and statistical analysis was performed in Excel.

Local maxima were used to count foci in ImageJ/FIJI, using the tool Find Maxima under the Process menu31. This tool requires a threshold input parameter (Prominence/Tolerance), where a threshold is set at the maximum value minus noise tolerance and local maxima must be greater than this threshold to be counted. To compare the effect of the input parameter, we performed measurements at Tolerance = 300 and 600. The final measurement was taken at the user-defined threshold where the user visually inspected different Tolerance values and chose a value that visually removed background (thresholds between 300 and 600) – a subjective process.

Imaris spots analysis were performed with Imaris 9.8 with background subtraction, using a spot size of 1 µm that was measured from the smallest distinguishable foci from the control image. For images with multiple nuclei, ImarisCell was used with the detection type “cell and vesicles”, where the nucleus in the DAPI channel was modelled as the “cell” and antibody-labeled foci were modelled as the “vesicles”. This enabled us to count the number of foci per nucleus for multiple cells. In each dataset, the automatic “quality” threshold was determined from the control image and the same threshold was used for every image in the dataset to be able to compare the outputs across datasets. Similar to the ICS analysis, only cells with an area between 400-2400 voxels were included for analysis.

Materials and Cell Culture

All chemicals were purchased from Sigma unless otherwise noted. All drugs were purchased from Selleck Chemicals. Cells were cultured in media supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin. OVCA429 were originally obtained from Dr. Michael Birrer, HT1080 were obtained from ATCC cultured in DMEM. SKOV3, HCC1937, HCC1395, MCF7, UWB1.289 and UWB1.289 +BRCA1 were obtained from ATCC and cultured in RPMI. UWB1.289res were made resistant to olaparib through long-term culture in up to 1 μM olaparib. Unless otherwise indicated, for MMS treatment, cells were incubated for 6 hours before fixation with 4% formaldehyde, whereas for treatment with PARP inhibitors and other drugs, cells were incubated for 24 hours before fixation.

Cell Dose Dependence Response

Cells were plated in a 96-well plate at 2×103 cells per well and treated with drug for 5 days in five replicate wells. Cell viability was then determined using Presto Blue (Thermo Fisher). Signal was averaged over replicate wells and normalized to blank (no cells) as well as to the levels of untreated wells. A sigmoidal response curve was fit to the average of two experiments in Prism.

Immunofluorescence and Fluorescence Microscopy

Cells were plated in 8-well slides (Fisher) or 384 well glass bottom plates (CellVis) to achieve 50% confluence overnight. Cells were then treated and fixed with 4% formaldehyde in PBS at room temperature for 10 minutes. Antibody staining was performed according to the manufacturers protocol (Cell Signaling Technologies or Abcam). Primary antibodies, that were vendor validated for specificity, were γH2AX (CST, #2577), RPA1 (CST, #2267), RAD51 (Abcam, 63801) and PCNA (CST, #2586), with Anti-rabbit IgG Fab2 Alexa Fluor 647 (CST, #4414) as the secondary fluorescent antibody. After antibody labeling, prolong gold with DAPI (CST) or DAPI (1 μg/ml) in PBS for well plate experiments was added and samples were stored for up to 1 week prior to imaging. Confocal imaging was performed on an Olympus FV1000 multiphoton/confocal microscope using an Olympus XLUMPlanFL N 20x objective, NA 1.00 with chromatic correction, pinhole at 1 Airy unit and 3x digital zoom. DAPI was excited with a 405 nm laser, while Alexa Fluor 647 was excited with a 635 nm laser. A multiband (DM 405/488/561/633 nm) dichroic mirror was used to direct the excitation laser and collect emission. Emission was then separated into two channels through a dichroic mirror (490nm) and DAPI signal was cleaned up by diffraction grating, which is the low band pass filter used by this instrument. Widefield imaging were performed on a Nikon Ti2 with a 20x (NA 0.75) objective using standard DAPI and Cy5 filter cubes.

Western Blotting

Cells were lysed with RIPA buffer (Thermo Fisher) containing protease inhibitors (Thermo Fisher), vigorously vortexed and incubated on ice for 15 minutes. Lysate was then pelleted by centrifugation at 14,000g for 15 min at 4 ºC. Supernatant was transferred to a clean tube and total protein was determined using a BCA assay (Pierce). Protein was loaded into a NuPAGE gel (Thermo Fisher) and transferred to nitrocellulose paper (Thermo Fisher). Blocking and antibody labeling was then carried out according to the antibody manufacturer’s protocol. GAPDH (R&D systems, #AF5718) was used as a loading control.

Statistics and Reproducibility

All experiments were performed in multiple repeats as indicated in the figure legends. Normalization to control was performed for conditions during each imaging session by dividing each datapoint by the average of the control data and the average control value was set to 1. For results that were collected from multiple different experiments the data were normalize to the control for each experiment. Statistical significance was calculated compared to the next lower dose using one way ANOVA or Students’ t test.

Results

Counting Foci via Fluctuation Analysis

Most foci counting methods are based on finding the local fluorescence intensity maxima either through intensity thresholding to create masks of high intensity single foci or implementing local maxima finding algorithms. Here, we evaluated traditional local maxima finding algorithms on the ovarian cancer cell line SKOV3 by creating resolvable DNA damage with the PARP inhibitor olaparib (10 µM) and antibody-labeling γH2AX to visualize foci (Fig. 1B). To enable clean isolation of nuclei, the cells were also labeled with DAPI (Fig. 1A). First, we used the Find Maxima tool in ImageJ/FIJI which detects local maxima to define individual foci. However, this approach relies heavily on the intensity threshold parameter (Prominence/Tolerance) used to remove the background signal. This parameter is defined by the user and is highly subjective – a high value causes some foci to be omitted from the count, yet, conversely, a low value leads to background signal being counted as foci. Therefore, we evaluated images at 2 different intensity thresholds (Fig. 1D-E) in 21 single cells to determine the ambiguity of the threshold parameter (Fig. 1G) before manually determining a threshold for each cell for the final measurement (Fig. 1H).

In addition, we implemented a different local maximum finding algorithm with automatic thresholding using Imaris Spots detection (Fig. 1F). Spots detection uses a Gaussian filter to smooth the image and foci are then located at the local maxima of the filtered image. Automatic thresholding is based on a k-means statistical method and Spots uses an expected foci size to identify multiple spots in a large cluster where individual foci are indistinguishable and could be classified as single foci with local maxima detection. Our results produce foci counts that are dependent on the approach used and setting implemented (Fig. 1 and S1). In our analysis, different thresholds/algorithms give different counts for the same cell due to different intensity foci (white arrows, Fig. 1I), different size foci (green arrows, Fig. 1I) and non-uniform background intensities (blue arrows, Fig. 1I).

We next employed fluctuation analysis via spatial ICS in the same set of images. For all ICS measurements immuno-staining was performed with primary and secondary antibodies based on the vendor protocol. No pre-extraction was performed prior to fixation. ICS was applied to individual, segmented nuclei in automated scripts. Uniquely ICS analyzes the entire dynamic range of detected fluorescence, not just intensity peaks that could be found with local maxima algorithms. We employed ICS analysis on segmented nuclei (Fig. 2D) and calculated the total number of particles (TNoP) parameter per nucleus for comparison to foci counting data from ImageJ/FIJI and Imaris. Here, TNoP is not a measure of foci, rather the amount of non-diffuse signal. Direct, cell by cell comparison of ICS TNoP with both the user-optimized threshold ImageJ Find Maxima (Fig. 2E) and Imaris Spots (Fig. 2F) foci counting results produced a strong correlation. Here, ICS is determining the total number of independent particles, leading to a TNoP value that is not directly counting resolvable foci. However, the correlation to the other approaches demonstates that ICS is measuring traditional foci forming DNA damage.

Particle Counting versus Aggregation in γH2AX signal

One major goal of foci counting in cells is to determine the degree of DNA damage. Therefore, after validating our foci counting methods at the individual cell level, we sought to evaluate ICS as a tool to quantify the DDR response in bulk populations of cells. We first treated SKOV3 cells with varying concentrations of methyl methane-sulfonate (MMS) or the PARP inhibitors veliparib or olaparib and evaluated DNA damage through immunofluorescence of γH2AX. MMS is a DNA alkylating agent that directly damages DNA by modifying guanine and adenine to instigate mispairing of bases and replication blocks32. PARP inhibitors, however, do not necessarily directly induce DNA damage, but stall the DNA damage response by “trapping” PARP1 and preventing recruitment of critical DNA damage response proteins23,33-35. Therefore, these two drug classes should have different impacts on focus formation and focus composition.

MMS treatment produces a wide range of γH2AX intensity and visible foci across individual cells (Fig. 3A-C; 0.1 mM MMS and Fig. S2; 0 mM and 1 mM MMS). Thus, analysis capable of detecting both low intensity foci and overlapping, highly abundant foci within a single experiment is needed for accurate measurements over a range of conditions. In the non-contrast adjusted γH2AX image (Fig. 3B), brighter γH2AX foci are clearly distinguishable, however once the image is contrast adjusted (Fig. 3C), lower intensity foci are apparent and the higher intensity foci appear as large foci clusters, especially in nuclei demonstrating more DNA damage. To overcome the inherent distribution of DNA damage across cells, we scaled up the analysis to examine hundreds of cells per image by acquiring large, stitched images (Fig. S3). Here, due to the dependency of selecting the proper background intensity per cell, we omitted the ImageJ Find Maxima analysis and analyzed data using ICS and Imaris Spots. To directly compare analyses, each measurement within an experiment was normalized to the average value of DMSO treated control cells.

Imaris Spots and ICS capture dose dependent DNA damage through γH2AX foci in SKOV3 cells.

(A-C) Confocal images of SKOV3 cells treated with 0.1 mM MMS; A) DAPI channel with 10 µm scale bar, B) γH2AX channel, C) γH2AX channel with intensity-adjustment. (D-L) Data for normalized intensity, normalized number of spots and normalized ICS DA of γH2AX signal across different MMS (D-F), veliparib (G-I) and olaparib (J-L) treatment concentrations in SKOV3 cells. Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. MMS; N=2423-3160, veliparib; N=1818-2922, olaparib; N=1489-2956. * indicates p<0.05, ** indicates p<0.001, *** indicates p<1×10−10 via one-way ANOVA significance testing.

In analyzing DNA damage, the amount of γH2AX per cell is represented by the normalized integrated fluorescence intensity (Fig. 3). We found that γH2AX intensity experienced small, significant increases at lower MMS doses but showed a massive (nearly 8X), significant increase at the highest dose, results that agree with SKOV3 sensitivity to MMS (Fig. S4). PARP inhibitor treatment produced similar results except for a slight decrease at 1 µM veliparib. The number of spots determined by Imaris captures the dose dependent DNA damage induced by MMS and PARP inhibitors as expected (Fig. 3), with the response to olaparib plateauing after 10 µM. Yet, surprisingly, at the highest dose of PARP inhibitor, the number of foci was higher under veliparib treatment, the weaker PARP inhibitor (Fig. S4).

We then analyzed the same images with ICS. Microscopic number density calculated by ICS often decreases in cases of significant aggregation, resulting in a smaller number of more densely packed clusters36,37, therefore we calculated the ICS degree of aggregation (DA) as a measurement that captures both low and high DNA damage. Critically, ICS DA is not a direct measurement of the number of foci, but rather a quantification of overall label clustering. DA results demonstrated both MMS and PARP inhibitor dose dependence (Fig. 3), similar to results obtained with Imaris Spots. Like Spots analysis, the ICS DA analysis of 100 µM veliparib produced a higher value than 100 µM olaparib, compared to their respective controls. Here, ICS DA proved comparable to Imaris Spots.

We next analyzed DNA damage dose dependence in another cell line, OVCA429. Fig. 4A-C displays OVCA429 cells exposed to 0.1 mM MMS, highlighting a range of γH2AX signal, spanning from only low diffuse signal in the nucleus, to single distinguishable γH2AX foci in the diffuse signal, to large clusters of different sizes in high intensity nuclei (Fig. 4C). γH2AX intensity significantly increases at 1 mM MMS (Fig. 4D) and Spots analysis captured the MMS dose dependent DNA damage (Fig. 4E) and ICS DA produced results similar to Spots analysis (Fig. 4F).

Imaris Spots and ICS DA capture dose dependent DNA damage through γH2AX in OVCA429 cells.

A-C) Confocal images of OVCA429 cells treated with 0.1 mM MMS; A) DAPI channel with 10 µm scale bar, B) γH2AX channel, C) γH2AX channel with intensity-adjustment. (D-L) Data for normalized intensity, normalized number of spots and normalized ICS DA of γH2AX signal across different MMS (D-F), veliparib (G-I) and olaparib (J-L) treatment concentrations in OVCA429 cells. Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. MMS; N=1194-2068, veliparib; N=1256-1997, olaparib; N=1445-2334. * indicates p<0.05, ** indicates p<0.001, *** indicates p<1×10−10 via one-way ANOVA significance testing.

Contrary to SKOV3 analysis, despite similar sensitivity to olaparib and veliparib (Fig. S4), γH2AX intensity in OVC429 in response to PARP inhibitors showed a lower change in signal with no significant change in intensity at higher concentrations (Fig. 4). However, Imaris Spots captured a significant dose-dependent response with all but 10 μM veliparib generating a significant increase from the next lower dose (Fig. 4). Thus, despite no changes in intensity, more foci are being formed at higher doses. ICS analysis, however, was more sensitive than Imaris Spots, capturing a significant dose-dependent change at each concentration (Fig. 4). Overall, in OVCA429, ICS DA produced similar results to Imaris Spots in olaparib treated cells, but measured dose dependent response in cells treated with veliparib where Imaris Spots was unable to capture the dose dependency.

ICS Captures Clustering of DNA Damage Proteins that do not form Resolvable Foci

Next, we sought to quantify the behavior of proteins that do not form resolvable foci during DNA damage without immunofluorescence pre-extraction, specifically Replication Protein A1 (RPA1) and RAD51. The RPA complex is required for major DNA repair pathways and modulates RAD51 recruitment to sites of DNA damage38,39. However, increased RPA1 and RAD51 concentrations created at sites of DNA repair are not always resolvable as individual foci39,40. In the absence of induced DNA damage, RPA1 immunofluorescence produces a diffuse signal throughout the nuclei in SKOV3 cells (Fig. 5). Only with contrast adjusting (Fig. 5C), does heterogeneous distribution of RPA1 become visible.

Dose-dependent DNA damage impact on RPA1 and RAD51.

(A-C) Confocal images of control SKOV3 cells, (D-F) SKOV3 cells treated with 1 mM MMS; (A&D) DAPI channel with 10 µm scale bar, (B&E) RPA1 channel, (C&F) RPA1 channel with intensity-adjustment. Normalized RPA1 Imaris Spots (G) and normalized ICS DA (H) in SKOV3 and OVCA429 cells (I&J). Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. N=1600-1872 for SKOV3 and N=1320-2198 for OVCA429. (K-M) Confocal images of control SKOV3 cells, (N-P) SKOV3 cells treated with 1 mM MMS; (K&N) DAPI channel with 10 µm scale bar, (L&O) RAD51 channel, (M&P) RAD51 channel with intensity-adjustment. Normalized RAD51 Imaris Spots (Q) and normalized ICS DA (R) in SKOV3 and OVCA429 cells (S&T). Shown are individual cells, average for each biological repeat (n=3, squares), and average with standard deviation. Results are normalized to average DMSO. N=1597-2241 for SKOV3 and N=1822-2311 for OVCA429. All results, * indicates p<0.05, ** indicates p<0.001, *** indicates p<1×10−10 via one-way ANOVA significance testing.

We used MMS to induce DNA damage in SKOV3 and OVCA429 cells. As expected, normalized intensity for RPA1 and RAD51 values (Fig. S5) did not display a dose dependence on MMS concentration. In SKOV3 cells treated with 1 mM MMS (Fig. 5), RPA1 recruitment to sites of DNA damage is observed alongside the diffuse signal of the nuclei. We first quantified the dose-dependent distribution of RPA1 with both Imaris Spots analysis (Fig. 5) and ICS DA measurements (Fig. 5). In SKOV3 cells, both Spots analysis (Fig. 5G) and ICS DA analysis (Fig. 5H) captured a dose dependent increase in RPA1 clustering, with the repeatability between experiments being much higher in ICS DA analysis suggesting that differences in immunofluorescence labeling between experiments impact Spots more than DA. In OVCA429 cells, Spots analysis demonstrated a more modest increase with increasing MMS dose (Fig. 5I), however, DA analysis produced a significant, larger increase in aggregation exceeding 2.5x of control at 1 mM MMS (Fig. 5J). Analysis of γH2AX clustering in OVCA429 cells treated with MMS (Fig. 4G) demonstrates a large increase at 1 mM MMS, suggesting that RPA1 DA analysis (Fig. 5J) is capturing DNA damage with a similar sensitivity to γH2AX analysis. The increased sensitivity in ICS DA versus Spots also indicates that measuring aggregation instead of foci counting could be more sensitive when analyzing proteins that do not form distinct foci.

Similar to RPA1, RAD51 does not form distinguishable foci in the nuclei in cells without pre-extraction (Fig. 5). With MMS treatment, Spots analysis of RAD51 foci captured an MMS dose dependency in OVCA429, but not SKOV3 cells (Fig. 5). However, ICS DA analysis captured dose dependent increases in aggregation (Fig. 5), with a 3X increase over control in SKOV3 at 1 mM MMS. Similar to RPA1 labeling experiments, the MMS dose dependence observed with RAD51 DA analysis is comparable to trends observed with γH2AX DA analysis for both cell lines. Thus, ICS DA proved to be more sensitive to RPA1 and RAD51 aggregation in response to DNA damage that induces significant γH2AX foci formation.

We then sought to determine if ICS analysis could provide insight into the DDR that had been previously unknown. Surprisingly, when we measured RPA1 DA induced by PARPi treatment we found different trends in different cell lines (Fig. 6). HCC1937 cells show a positive dependency of RPA1 DA with olaparib concentration. However, HT1080 cells show the opposite relationship, with RPA1 DA decreasing with increasing olaparib concentration. We then expanded our analysis to more cell lines and measured RPA1 DA, normalized to control, in response to 1 μM treatment of three different PARPi (Fig. 6B). Plotting the RPA1 DA value for each PARPi as a function of the cellular sensitivity (cell IC50) to each drug generated a curious pattern (Fig. 6C). Here, cell lines that were most sensitive to PARPi treatment demonstrated a decreasing RPA1 DA with drug potency (inverse of cell IC50), while cell lines that were more resistance had the opposite response. Fitting a linear curve to the RPA1 DA vs. cell IC50 data for each cell line and plotting the slope of that fit against the average cell line sensitivity to PARPi treatment produced a striking pattern (Fig. 6D). These results suggest the capacity to recruit RPA1 into the DDR imparts resistance to PARPi treatment. Surprisingly, the expression level of RPA1 in each cell line had no correlation to cell line sensitivity to PARPi (Fig. 6E&F). Therefore, these results demonstrate the power of ICS to identify DDR responses that would be otherwise very difficult to detect.

RPA1 recruitment correlates with cell sensitivity to PARPi.

(A) RPA1 DA as a function of olaparib concentration for HT1080 and HCC1937 cells, shown are average with SEM, n >= 504 cells, 3 biological repeats, and linear fit. (B) RPA1 DA in cells treated with 1 μM PARPi overnight, shown are average, normalized to untreated control, with SEM, n >= 350 cells, 2 biological repeats. (C) RPA1 DA as a function of cell line IC50 for each cell line and PARPi shown in Fig. 5b. Shown are average, normalized to untreated control, with SEM, n >= 350 cells, 2 biological repeats and linear fit (prism) for each cell line. (D) The linear fit slope of RPA DA vs. cell line IC50 as a function of average PARPi z-score for each cell line. Shown are slope fit with SEM and average z-score over three PARPi with st. dev. (E) Western blot of RPA1 expression levels for each cell line, with GAPDH loading control. (F) Normalized protein expression level in each cell line as a function of PARPi z score. Shown are average z-score over three PARPi with st. dev.

ICS can be Applied to a Diverse Set of Imaging and DNA Damage Response Measurements

Since ICS analysis is used on standard fluorescence microscopy images and we obtain a DAPI image every experiment we sought to incorporate cell cycle classification into measurements. Here, to demonstrate the broad applicability of ICS we performed measurements in 384 well plates on a widefield microscope. Previous results have established that DAPI intensity can be used to stage cells41. We confirmed our ability to distinguish cell stage through DAPI intensity by treating SKOV3 cells with palbociclib, camptothecin or etoposide, which cause cells to stall at different cell cycle stages. Similar to previous findings41, we found that palbociclib stalled cells in G1, camptothecin stalled cells in S and etoposide stalled cells in G2 (Fig. 7A). Once confirmed, we set DAPI intensity thresholds for each cell cycle stage. Thresholds were established automatically for each experiment by finding the peak of the DAPI intensity histogram of all control cells. We then classified G1 phase as cells with DAPI intensities within 10% of the DAPI intensity at the histogram peak, S phase as cells with DAPI intensity within 15% of the histogram peak times 1.5, and G2 phase as cells with DAPI intensity within 20% of the histogram peak times 2 (Fig. 7B).

Incorporating and validating cell cycle data in widefield imaging.

(A) Histograms of DAPI intensities for cells treated with palbociclib, camptothecin or etoposide, 1 μM overnight. (B) Representative assignment of cell cycle based on DAPI histogram (grey), using the first peak as G1 cells. Normalized PCNA intensity (C) and DA (D) in SKOV3 cells that had been classified by DAPI intensity into phases of the cell cycle. Normalized γH2AX intensity (E) and DA (F) in SKOV3 cells that had been classified by DAPI intensity into phases of the cell cycle. Shown are average n>200 cells per classification and standard deviation.

We validated our cell cycle staging approach through immunofluorescence of γH2AX and PCNA. Untreated SKOV3 and HCC1937 cells were classified by cell cycle stage and both ICS DA and intensity were calculated for the two labels. Data were normalized to unclassified average. We found that PCNA intensity increased from G1 to S, which was similar to G2 for both cell lines (Fig. 7C), suggesting that PCNA expression levels increase during S phase and plateau. However, PCNA DA increased from G1 to S, then decreased from S to G2 for both cell lines (Fig. 7D), suggesting that PCNA has the highest aggregation in S phase, as would be expected. PCNA is recruited to DNA during synthesis, which will increase the overall aggregation as measured here. We found that γH2AX intensity increased from G1 to S, then plateaued in G2 for HCC1937 cells but increased from S to G2 in SKOV3 cells (Fig. 7E). γH2AX DA increased from G1 to S and from S to G2 for both cell lines (Fig. 7F). These γH2AX results confirm that γH2AX is more prominent in S and G2 than G1.

We then measured γH2AX DA response in SKOV3 cells treated with DNA damaging agents olaparib, bleomycin or etoposide (Fig. 8). Cells were classified by cell cycle and the data aggregated and plotted as a function of drug concentration. Here we observed significant changes in both γH2AX DA and intensity as a function of damaging agent concentration. However, the γH2AX DA plateaued at higher doses, while the γH2AX intensity had a consistent dose dependent increase. These measurements suggest that the amount of γH2AX in the cell is a function of the degree of DNA damage, but the clustering of γH2AX reaches a peak at very large amounts of damage. Interestingly, we did not observe large differences between cell cycle stages for any of the compounds. Overall, olaparib and etoposide generated more γH2AX intensity in S and G2 phases while bleomycin generated the highest γH2AX intensity at 10 μM in G1 cells. However, the number of cells in each cell cycle was very dependent on the drug and concentration, with all compounds inducing more cells to stall at G2.

Cell cycle analysis of DDR ICS in widefield.

Normalized γH2AX DA and intensity of SKOV3 cells treated with DMSO or olaparib (A), bleomycin (B) or etoposide (C) at 3 different doses (in μM) for 24 hours and separated by cell cycle stage based on DAPI intensity. Data were normalized to cell cycle independent DMSO average. Also shown are the ratios of the number of cells in G2 to G1 for each treatment condition. Data are n=7 biological repeats (box = average) with single cell values and standard deviation, n>200 cells per condition. For clarity all data are significantly different, p<0.05 Student’s t test from the next lower concentration unless marked – ns = no significance, dec = significant decrease.

Discussion

Fundamentally, DDR foci are resolvable when the labeled protein concentration difference between foci and the rest of the nucleus is high enough to distinguish in fluorescence microscopy. The local concentration at sites of DNA damage is driven by either protein recruitment, or, in the case of γH2AX, post translational modifications42. Yet, most DDR proteins do not cluster in DNA damage at concentrations high enough to resolve recruitment in the absence of massive, artificial DNA damage. γH2AX is the most prominent marker of DNA damage foci largely because H2AX phosphorylation occurs at sites of double stranded breaks producing a marker with very low background signal. However, recent super resolution results identify γH2AX foci as groups of smaller foci22, suggesting that not all foci are equal, and large foci in diffraction-limited microscopy represent a greater degree of DNA damage than smaller foci. Therefore, foci count alone may not accurately represent DNA damage.

Clustering analysis with ICS overcomes both limits of counting traditional markers and measuring non-traditional markers as metrics of DNA damage or to study their role in the DDR. Initially we used the TNoP parameter to match foci counts in individual cells from two methods, however ICS calculated degree of aggregation (DA) captures overall clustering of protein and is therefore the more appropriate ICS measurement. Using ICS DA, we were able to detect the dose dependent response to two PARP inhibitors, matching or exceeding dose-dependent sensitivity of foci counting via Imaris Spots. Furthermore, ICS DA was uniquely able to measure dose-dependent DNA damage using RPA1 and RAD51 as DNA damage markers. The DNA damage response of RPA1 and RAD51 measured through immuno-staining have previously employed pre-extraction38,43-45, here the use of ICS removes this subjective sample preparation step to simplify measurements of the DDR activity of these proteins. The accuracy of these measurements was validated by the similarity of the dose dependency to γH2AX measurements. γH2AX is the most common marker used to define DDR foci. Therefore, ICS is able to evaluate DNA damage with non-traditional markers or evaluate the recruitment of proteins to DNA damage.

ICS analysis of γH2AX in OVCA429 cells treated with PARP inhibitors, where there was not a dose dependent increase of γH2AX intensity, revealed an increase in DA, indicating larger foci cluster formation was a merging of smaller clusters to produce an overall decrease in the number of foci clusters36,37. The formation of large foci can impact foci counting methods, however, Imaris Spots uses a spot size parameter that aids in placing multiple smaller spots per large foci. Yet, this spot size parameter failed to consistently capture the dose dependent increases in DNA damage where γH2AX intensity was not also increased. Spatial ICS is well suited to measure large aggregations and has previously been used to count distinguishable objects larger than the diffraction limit such as fluorescent beads28 and dendritic spines46, by using an intensity threshold to remove the background signal. This can be supplemented by various image analysis algorithms such as Gaussian filters to smooth out fluctuations in the objects so they can be detected as one object. However, we did not need an intensity threshold to remove the background signal, and only used a background subtraction that represented the intensity in a part of the image with no cells to account for autofluorescence outside of the cells.

We then used ICS to assess the involvement of RPA1 in the DDR, with results suggesting that recruitment correlates with cellular sensitivity to PARPi. These results were not apparent when measure RPA1 expression levels across cell lines, a typical evaluation of cell line sensitivity, indicating that ICS is capable of revealing underlying biology not detected by traditional approaches. We also demonstrated that cell cycle analysis can be readily performed on results to classify cells by cell stage for measurements of the impact of cellular stage on the DNA damage response. As a simple image analysis technique, ICS can be applied broadly to incorporate other analysis as well.

Here, we demonstrated that ICS is an alternative technique to foci counting, where the clustering involved in protein recruitment during DNA damage can be fully captured on the molecular level. ICS measures the degree of clustering and considers the low intensity signal contributions that are often lost during foci segmentation. Our findings correlate with prior studies characterizing foci formation/clustering and the DNA damage response using advanced optical techniques such laser micro-irradiation47 and super resolution microscopy22. Advantageously, ICS can be implemented on any fluorescence image as long as the square relative fluorescence intensity fluctuations are detectable above noise fluctuations. ICS does not require laser micro-irradiation to induce detectable clustering23,47,48. Also, it overcomes the complications introduced when using pre-extraction in immunofluorescence49, where soluble protein is removed through permeabilizing the cell for an arbitrary amount of time, likely removing some recruited protein as well. Thus, ICS enables simple, standard immunofluorescence labeling techniques. Other approaches that increase image resolution, such as super-resolution microscopy, will enhance the identification of single molecules and could be used to detect dimers with enough resolution enhancement. However, the definition of clustering would remain unclear as boundaries for each cluster (or foci) would have to be established, which requires subjective interpretation of the image. The overall approach, using a combination of conventional fluorescence microscopy, antibody labeling, and ICS analysis, provides a molecular level understanding for characterizing protein recruitment and signaling in a variety of applications from capturing the DNA damage response to evaluating cancer therapies. We expect adoption of this approach will both lead to a more objective measure of DNA damage and provide a tool to evaluate the role of every DDR protein during the DNA damage response.